---
title: "Data Extraction from the MIMIC-III Database"
author: "<h3>Brandon Cummings</h3>"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: default
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

# Introduction

This markdown notebook is intended to house the code used to generate the data stored in the `./data/` directory. It calls utilities found in `./utilities/data_extraction_tools.R` to make a connection to the PostgreSQL database, then extract requisite data. Running the notebook will overwrite data in the above directory.

**Import functionality**
```{r message=F, warning=F}
source('utilities/data_extraction_tools.R')
con <- open.connection()
```

In several cases, entire tables will be loaded and saved. This includes dictionary tables such as `D_ICD_DIAGNOSES` table, as well as smaller tables such as `DIAGNOSES_ICD` which can be used to generate patient cohorts.

# Extracting whole tables

```{r message=F, warning=F}
# D_ICD_DIAGNOSES
d_icd_diagnoses <- get.table(con, 'D_ICD_DIAGNOSES')
save(d_icd_diagnoses, file='./data/d_icd_diagnoses.RData')

# DIAGNOSES_ICD
diagnoses_icd <- get.table(con, 'DIAGNOSES_ICD')
save(diagnoses_icd, file='./data/diagnoses_icd.RData')

# ADMISSIONS
admissions <- get.table(con, 'ADMISSIONS')
save(diagnoses_icd, file='./data/admissions.RData')

```
