---
title: "EFFECT_OPTIMIE_OPTIMIZE"
author: "<h3><p>MIMIC-III SOCR Team</p></h3>"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: default
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
subtitle: <h2><u>Winter 2019, SOCR-MDP</u></h2>
---

**Libraries Used:**
```{r message=F, warning=F}
# Data I/O
#library('RPostgreSQL')    # access MIMIC-III
library('reshape2')       # melting dataframes
library('dplyr')          # misc. organization
library('bigrquery')     # Query data from Google BigQuery
# Data preparation
library('psych')          # descriptive stats

# Plots and tables
library('knitr')          # knitting Rmd to HTML; kable() function
library('kableExtra')     # extra formating options for knitr tables
library('ggplot2')        # 2d plotting
library('ggpubr')         # extra formatting options for ggplot
library('gtools')
library("lubridate")


project = 'mimic-iii-244018'
```

# Inital Cohort
Patients with CHF who died outside of the hospital. 

```{r eval=T, message=F, warning=F}
# read in cohort from CSV (includes DOD)
# also already includes SBP and Creatinine
# cohort_data <- read.csv(file="cohort_with_lab.csv", header = T)

# kable(head(cohort_data), caption="Sample of cohort data") %>%
 # kable_styling(bootstrap_options='striped')
```

# Discussion of EFFECT, OPTIME, OPTIMIZE

## EFFECT Scoring System
The Enhanced Feedback for Effective Cardiac Treatment (EFFECT) scoring system is a derivation and validation of a clinical model that aims to predict mortality among patients hospitalized for heart failure. The objective is to identify high-risk heart failure patients at hospital discharge, thus allowing more effective triage to management strategies. In order to apply the EFFECT scoring system on the MIMIC-III dataset, we retrieved the cohort patients' age, systolic blood pressure, elevated blood urea nitrogen level, sodium concentration, cerebrovascular disease, hepatic cirrhosis, history of cancer and hemoglobin level. {r} [(M. O'Connor 2010)] {https://www.sciencedirect.com/science/article/pii/S0735109709040595}

## OPTIME Scoring System
The Outcomes of a Prospective Trial of Intravenous Milrinone for Exacerbations of Chronic Heart Failure (OPTIME-CHF) trial aims to evaluate the treatment strategies in the care of population with acute exacerbations of chronic heart failure. The primary goal of this scoring system is a reduction in the total hospital days for cardiovascular events within 60 days after therapy. In order to apply the OPTIME-CHF scoring system on the MIMIC-III dataset, we retrieved the cohort patients' age, elevated blood urea nitrogen level and systolic blood pressure. {r} [(Duke Clinical Research Institue 2004)] {https://www.ncbi.nlm.nih.gov/pubmed/15599835}

## OPTIMIZE Scoring System
The Organized Program to Initiate Lifesaving Treatment in Hospitalized Patients with Heart Failure (OPTIMIZE-HF) is a comprehensive hospital-based registry and performance-improvement program designed to predict in-hospital mortality in patients hospitalized for heart failure. In order to apply the OPTIMIZE-HF scoring system on the MIMIC-III dataset, we retrieved the cohort patients' age, weight, systolic blood pressure, creatinine level, liver disease, history of depression and reactive airway disease. {r} [(Duke Clinical Research Institue 2008)] {https://www.ncbi.nlm.nih.gov/pubmed/18926148}


## Read in cohort
```{r}
cohort <- read.csv("new_cohort.csv")
```

## Extract itemcode 

### BUN item codes
```{r}
sql = "SELECT ITEMID FROM [physionet-data.mimiciii_clinical.d_items] 
        WHERE LABEL LIKE \"%BUN%\""
bun.itemcodes = query_exec(sql, project=project)
bun.itemcodes = bun.itemcodes$ITEMID
```
 
### Creatinine item code
```{r}
sql = "SELECT ITEMID FROM [physionet-data.mimiciii_clinical.d_items] 
        WHERE LABEL LIKE \"%creatinine%\" OR LABEL LIKE \"%Creatinine%\" "
creatinine.itemcodes = query_exec(sql, project=project)
creatinine.itemcodes = creatinine.itemcodes$ITEMID
```

### Hemoglobin item codes
```{r}
sql = "SELECT ITEMID FROM [physionet-data.mimiciii_clinical.d_items] 
        WHERE LABEL LIKE \"%hemoglobin%\" OR LABEL LIKE \"%Hemoglobin%\" "
hemoglobin.itemcodes = query_exec(sql, project=project)
hemoglobin.itemcodes = hemoglobin.itemcodes$ITEMID
```

### Sodium itemcodes
```{r}
sql = "SELECT ITEMID FROM [physionet-data.mimiciii_clinical.d_items] 
        WHERE LABEL LIKE \"%sodium%\" OR LABEL LIKE \"%Sodium%\" "
sodium.itemcodes = query_exec(sql, project=project)
sodium.itemcodes = sodium.itemcodes$ITEMID
```

## Extract lab results

This pull.last.event function is going to be used when extracting lab results
```{r message=F, warning=F}
pull.last.event <- function(hadm_id, itemcodes, return.fields, table, project, max_pages=Inf) {
  sql <- sprintf("
                 SELECT *
                 FROM [physionet-data.mimiciii_clinical.%s] AS table1
                 INNER JOIN (
                   SELECT hadm_id, MAX(charttime), MAX(row_id)
                   FROM [physionet-data.mimiciii_clinical.%s]
                   WHERE itemid IN (%s)
                   AND hadm_id IN (%s)
                   GROUP BY hadm_id
                 ) AS table2
                 ON table1.row_id = table2.f1_",
                 table, table, 
                 paste(itemcodes, collapse=", "),
                 paste(hadm_id, collapse=", ")
  )
  
  data <- query_exec(sql, project=project, max_pages=max_pages)
  colnames(data) <- gsub('table[0-9]_', '', colnames(data))
  return(data[ , return.fields])
}
```

### Systolic Blood Pressure

```{r message=F, warning=F}
sbp.itemcodes <- c( 6, 51, 442, 3313, 3315, 3317, 3321, 3323, 3325, 6701, 228152, 224167, 227243, 220050, 220179, 225309 )
return.fields <- c('HADM_ID', 'VALUENUM')
data.sbp <- pull.last.event(cohort$HADM_ID, sbp.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.sbp)[2] = "SBP"
```

### BUN

```{r}
return.fields <- c('HADM_ID', 'VALUENUM')
data.bun <- pull.last.event(cohort$HADM_ID, bun.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.bun)[2] = "BUN"
```

### Creatinine

```{r}
return.fields <- c('HADM_ID', 'VALUENUM')
data.creatinine <- pull.last.event(cohort$HADM_ID, creatinine.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.creatinine)[2] = "Creatinine"
```

### Weight

```{r}
weight.itemcodes <- c(226531, 763)
return.fields <- c('HADM_ID', 'VALUENUM')
data.weight <- pull.last.event(cohort$HADM_ID, weight.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.weight)[2] = "Weight"
```

### Sodium

```{r}
sodium.itemcodes <- c(220645)
return.fields <- c('HADM_ID', 'VALUENUM')
data.sodium <- pull.last.event(cohort$HADM_ID, sodium.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.sodium)[2] = "Sodium"
```

### Hepatic Cirrhosis

[5715 is the icd9 code for hepatic cirrhosis](http://www.icd9data.com/2012/Volume1/520-579/570-579/571/571.5.htm)

```{r}
sql <- "SELECT hadm_id
        FROM [physionet-data.mimiciii_clinical.diagnoses_icd]
        WHERE icd9_code = \"5715\" "

cirr.cohort <- query_exec(sql, project=project)

```
## Combine the data

```{r}
cohort.with.lab <- left_join(cohort, data.sbp, by='HADM_ID') %>%
                left_join(., data.creatinine, by='HADM_ID') %>%
                left_join(., data.weight, by='HADM_ID') %>%
                left_join(., data.sodium, by='HADM_ID')
```

```{r}
cohort.with.lab$cirr = FALSE
cohort.with.lab$cirr = cohort.with.lab$HADM_ID %in% cirr.cohort$hadm_id
# write.csv(cohort.with.lab, "cohort_with_lab.csv")
```

## Liver Disease
```{r eval=T, message=F, warning=F}
# History of Liver Disease
sql_icd9_liver <- sprintf("
                    SELECT SUBJECT_ID
                    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
                    WHERE icd9_code IN ('570', '5710', '5712', '5713', '5715', '5728', '5807', '5730', '5738', '5739', '5718', '5719') ")
ld_data <- query_exec(sql_icd9_liver, project, max_pages = 1)
ld_vector <- pull(ld_data, SUBJECT_ID) # subject ids for history of liver disease

cohort_data$Liver <- rep(0, nrow(cohort_data))
cohort_data$Liver <- ifelse(cohort_data$SUBJECT_ID %in% ld_vector, 1, 0)
```

# History of Depression
We referenced [Appendix A.2 from Healthcare Cost and Utilization Project (HCUP) Statistical Briefs](https://www.ncbi.nlm.nih.gov/books/NBK409512/table/sb216.t6/) in selection of ICD9 codes related to depression. 
```{r eval=T, message=F, warning=F}
sql_icd9_depression <- sprintf("
    SELECT SUBJECT_ID
    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
    WHERE icd9_code IN ('29021', '29013', '30112', '29621', '29682', '311', '2980', '29633', '29632', '29626', '29625', '29624', '29623', '29620', '29622', '3091', '29634', '29636', '29635', '29631', '29630') ") 

depression_data <- query_exec(sql_icd9_depression, project, max_pages = 1)
depression_vector <- pull(depression_data, SUBJECT_ID) # subject ids of all patients who ever had depression
                
cohort_data$Depression <- rep(0,nrow(cohort_data))
cohort_data$Depression <- ifelse(cohort_data$SUBJECT_ID %in% depression_vector, 1, 0)
```

# History of Reactive Airway Disease
```{r eval=T, message=F, warning=F}
sql_icd9_reac <- sprintf("
                    SELECT SUBJECT_ID
                    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
                    WHERE icd9_code IN ('49390', '49310', '49320', '49311', '49312', '49321', '49322') ")
reac_data <- query_exec(sql_icd9_reac, project, max_pages = 1)
reac_vector <- pull(reac_data, SUBJECT_ID)

cohort_data$Reactive_airway <- rep(0, nrow(cohort_data))
cohort_data$Reactive_airway <- ifelse(cohort_data$SUBJECT_ID %in% reac_vector, 1, 0)
```

# History of Cancer
We referenced [Weiner, M.D. et. al’s paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1480106/?fbclid=IwAR2rOQKA9k-61IqwWto3c_6mOcNfdBUzbd2R6y0h6JhCoJTaJ3LynO9B2Fw) for which ICD9 codes to extract for a subject’s history of cancer. We used all groups of cancer so to include anyone with a history of cancer. 
```{r eval=T, message=F, warning=F}
c_icd <- c(seq(1400, 1649, by=1), seq(1700, 1759, by=1), seq(1800, 1809, by = 1), 
                             seq(1820, 1969, by=1), seq(1990, 1999, by=1), seq(2350, 2399, by=1))
c_string_icd <- toString(sprintf("'%s'", c_icd))

sql_icd9_cancer <- sprintf("
SELECT SUBJECT_ID
    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
    WHERE icd9_code IN (%s)", paste(c_string_icd, collapse=", "))
cancer_data <- query_exec(sql_icd9_cancer, project, max_pages = 1)
cancer_vector <- pull(cancer_data, SUBJECT_ID)

cohort_data$Cancer <- rep(0, nrow(cohort_data))
cohort_data$Cancer <- ifelse(cohort_data$SUBJECT_ID %in% cancer_vector, 1, 0)


kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# OPTIMIZE read_in features - <feature> #WILL NEED TO HAVE feature extraction code
will fix to put in extraction code

```{r eval=T, message=F, warning=F}
# read in other OPTIMIZE features and add to cohort

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# EFFECT feature extraction - Depression, Cancer

```{r eval=T, message=F, warning=F}
kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# EFFECT feature read in - 

```{r eval=T, message=F, warning=F}
kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# OPTIMIE features left to get - 

```{r eval=T, message=F, warning=F}
# OPTIMIZE features

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# Imputation of missing values and related plots
add references to dinov's tutorial

```{r eval=T, message=F, warning=F}
library(Amelia)
sim_data.df <- read.csv(file='DATA_MASTER.csv', header=TRUE)
dim(sim_data.df)
amelia.out <- amelia(sim_data.df, m = 5)
amelia.out
amelia.imputed.5 <- amelia.out$imputations[[5]]

```

# EFFECT Score calculation! and basic summary stats?
```{r eval=T, message=F, warning=F}
# EFFECT
#age, cbv disease, dementia, COPD, BUN + cirrhosis + hemoglobin + resp, SBP, SODIUM
cohort_data$EFFECT <- rep(0, nrow(cohort_data))
for (i in 1:nrow(cohort_data)){
  score <- 0
  #age
  score <- score + cohort_data$AGE[i]
  if( cohort_data$CV_Disease[i] == 1){
    score <- score + 10
  }
  if( cohort_data$Dementia[i] == 1){
    score <- score + 10
  }
  if (cohort_data$COPD[i] == 1){
    score <- score + 20
  }
  # do nothing for hemoglobin i guess
  if(cohort_data$BUN[i] <= 60){
    score <- score + cohort_data$BUN[i]
  }
  if(cohort_data$RESP_RATE[i] >= 20 && cohort_data$RESP_RATE[i] <= 45){
    score <- score + cohort_data$RESP_RATE[i]
  }
  if( cohort_data$cirrhosis[i] == 1){
    score <- score + 25
  }
  # sbp
  if(cohort_data$SBP[i] >= 180){
    score <- score - 60
  } else if(cohort_data$SBP[i] >= 160){
    score <- score - 55
  } else if(cohort_data$SBP[i] >= 140){
    score <- score - 50
  } else if(cohort_data$SBP[i] >= 120){
    score <- score- 45
  } else if(cohort_data$SBP[i] >= 100){
    score <- score - 40
  } else if(cohort_data$SBP[i] >= 90){
    score <- score - 35
  } else if(cohort_data$SBP[i] < 90){
    score <- score - 30
  }
  if(cohort_data$SODIUM[i] < 136){
   score <- score + 10
  }
  cohort_data$EFFECT[i] <- score
}
```

# OPTIME Score
```{r eval=T, message=F, warning=F}
# OPTIME
cohort_data$OPTIME <- rep(0, nrow(cohort_data))
for (i in 1:nrow(cohort_data)){
  score = 0
  if(is.na(AGE)){
  }else if(AGE > 20 && AGE <= 30){
    score = score + 8
  }else if(AGE > 30 && AGE <= 40){
    score = score + 17
  }else if(AGE > 40 && AGE <= 50){
    score = score + 25
  }else if(AGE > 50 && AGE <= 60){
    score = score + 33
  }else if(AGE > 60 && AGE <= 70){
    score = score + 41
  }else if(AGE > 70 && AGE <= 80){
    score = score + 50
  }else if(AGE > 80 && AGE <= 90){
    score = score + 58
  }
  
  if(is.na(sodium)){
  }else if(sodium < 115){
    score = score + 79
  }else if(sodium < 120){
    score = score + 69
  }else if(sodium < 125){
    score = score + 59
  }else if(sodium < 130){
    score = score + 49
  }else if(sodium < 135){
    score = score + 30
  }else if(sodium < 140){
    score = score + 20
  }else if(sodium < 145){
    score = score + 10
  }
  
  if(is.na(SBP)){
  }else if(SBP < 80){
    score = score + 94
  }else if(SBP < 90){
    score = score + 86
  }else if(SBP < 100){
    score = score + 77
  }else if(SBP < 110){
    score = score + 69
  }else if(SBP < 120){
    score = score + 60
  }else if(SBP < 130){
    score = score + 51
  }else if(SBP < 140){
    score = score + 43
  }else if(SBP < 150){
    score = score + 34
  }else if(SBP < 160){
    score = score + 26
  }else if(SBP < 170){
    score = score + 17
  }else if(SBP < 180){
    score = score + 9
  }
  
  if(is.na(BUN)){
  }else if(BUN < 5){
    score = score + 10
  }else if(BUN < 10){
    score = score + 20
  }else if(BUN < 15){
    score = score + 30
  }else if(BUN < 20){
    score = score + 40
  }else if(BUN < 25){
    score = score + 50
  }else if(BUN < 30){
    score = score + 60
  }else if(BUN < 35){
    score = score + 70
  }else if(BUN < 40){
    score = score + 80
  }else if(BUN < 45){
    score = score + 90
  }else{
    score = score + 100
  }
  
  cohort_data$OPTIME[i] <- score
}


```

# OPTIMIZE Score
```{r eval=T, message=F, warning=F}
# OPTIMIZE
#age, cbv disease, dementia, COPD, BUN + cirrhosis + hemoglobin + resp, SBP, SODIUM
#cohort_data_complete <- read.csv('DATA.csv')

cohort_data$OPTIMIZE <- rep(0, nrow(cohort_data))
for (i in 1:nrow(cohort_data)){
  score <- 0
  #age
  if( cohort_data$AGE[i] < 30){
    score <- score + 0
  } else if( cohort_data$AGE[i] < 40){
    score <- score + 2
  }  else if( cohort_data$AGE[i] < 50){
    score <- score + 5
  }  else if( cohort_data$AGE[i] < 60){
    score <- score + 7
  }  else if( cohort_data$AGE[i] < 70){
    score <- score + 10
  }  else if( cohort_data$AGE[i] < 80){
    score <- score + 12
  }  else if( cohort_data$AGE[i] < 90){
    score <- score + 15 
  } else if( cohort_data$AGE[i] >= 90){
    score <- score + 17 
  }  
  # weight, kg
  if( cohort_data$WEIGHT[i] < 70){
    score <- score + 9
  } else if( cohort_data$WEIGHT[i] < 90){
    score <- score + 7
  } else if( cohort_data$WEIGHT[i] < 110){
    score <- score + 5
  } else if( cohort_data$WEIGHT[i] < 130){
    score <- score + 3
  } else if( cohort_data$WEIGHT[i] >= 130){
    score <- score + 2
  }
  # sbp
  if(cohort_data$SBP[i] < 90){
    score <- score + 24
  } else if(cohort_data$SBP[i] < 110){
    score <- score + 20
  } else if(cohort_data$SBP[i] < 130){
    score <- score + 17
  } else if(cohort_data$SBP[i] < 150){
    score <- score + 13
  } else if(cohort_data$SBP[i] < 170){
    score <- score + 11
  } else if(cohort_data$SBP[i] < 190){
    score <- score + 9
  } else if(cohort_data$SBP[i] < 210){
    score <- score +8
  } else if(cohort_data$SBP[i] < 230){
    score <- score +6
  } else if(cohort_data$SBP[i] < 250){
    score <- score +4
  } else if(cohort_data$SBP[i] < 270){
    score <- score +2
  } else if(cohort_data$SBP[i] >= 270){
    score <- score + 0
  } 
  
  #sodium
  if(cohort_data$SODIUM[i] < 112.5){
   score <- score + 12
  } else if(cohort_data$SODIUM[i] < 117.5){
   score <- score + 10
  } else if(cohort_data$SODIUM[i] < 122.5){
   score <- score + 8
  } else if(cohort_data$SODIUM[i] < 127.5){
   score <- score + 6
  } else if(cohort_data$SODIUM[i] < 132.5){
   score <- score + 4
  } else if(cohort_data$SODIUM[i] < 137.5){
   score <- score + 2
  } else if(cohort_data$SODIUM[i] >= 137.5 ){
   score <- score + 0
  }
  #creatinine 
  if (cohort_data$CREATININE[i] < 0.5){
    score <- score + 0
  } else if (cohort_data$CREATININE[i] < 1.5){
    score <- score + 5
  } else if (cohort_data$CREATININE[i] < 2.5){
    score <- score + 9
  } else if (cohort_data$CREATININE[i] < 3.5){
    score <- score + 14
  } else if (cohort_data$CREATININE[i] >= 3.5){
    score <- score + 19
  }
  
  # history of liver disease 
  if(cohort_data$Liver[i] == 1){
    score <- score + 8
  }
  
  #history of reactive airway disease 
  if( cohort_data$Reactive_airway[i] == 1){
    score <- score + 4
  }
  #history of depression
  if( cohort_data$Depression[i] == 1){
    score <- score + 4
  }
  
  cohort_data$OPTIMIZE[i] <- score
}
```

# Binning patients based on time til death outside of hospital


```{r eval=T, message=F, warning=F}
# Death binning code
# Emily has it and will add if we want it

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# Plots analyzing binned death for scores 
analysis of different scores motives (optimize looking at 6-9 month mortality, ect)


```{r eval=T, message=F, warning=F}
# plots of binned deaths!
```

# Mortality as continuous variable plots and analysis
add references to dinov's tutorial
