---
title: "EFFECT_OPTIMIE_OPTIMIZE"
author: "<h3><p>MIMIC-III SOCR Team</p></h3>"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: default
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
subtitle: <h2><u>Winter 2019, SOCR-MDP</u></h2>
---

**Libraries Used:**
```{r message=F, warning=F}
# Data I/O
#library('RPostgreSQL')    # access MIMIC-III
library('reshape2')       # melting dataframes
library('dplyr')          # misc. organization
library('bigrquery')     # Query data from Google BigQuery
# Data preparation
library('psych')          # descriptive stats

# Plots and tables
library('knitr')          # knitting Rmd to HTML; kable() function
library('kableExtra')     # extra formating options for knitr tables
library('ggplot2')        # 2d plotting
library('ggpubr')         # extra formatting options for ggplot
library('gtools')
library("lubridate")


project = 'wallems-mimic'
```

# Inital Cohort
Patients with CHF who died outside of the hospital. 

```{r eval=T, message=F, warning=F}
# read in cohort from CSV (includes DOD)
# also already includes SBP and Creatinine
cohort_data <- read.csv(file="cohort_with_lab.csv", header = T)

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# Discussion of EFFECT, OPTIME, OPTIMIZE
a little about the scores, including what they are predicting (like 6-9month mortality?) 
table of how score calculated
with references!

# OPTIMIZE feature extraction - Liver Disease, Reactive Airway Disease, other things
each feature probs needs to be it's own box so that we can talk about how choose icds's and such
i think. or talk about how choose each feature all together and then just stick queries and code in one block?

# Liver Disease
```{r eval=T, message=F, warning=F}
# History of Liver Disease
sql_icd9_liver <- sprintf("
                    SELECT SUBJECT_ID
                    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
                    WHERE icd9_code IN ('570', '5710', '5712', '5713', '5715', '5728', '5807', '5730', '5738', '5739', '5718', '5719') ")
ld_data <- query_exec(sql_icd9_liver, project, max_pages = 1)
ld_vector <- pull(ld_data, SUBJECT_ID) # subject ids for history of liver disease

cohort_data$Liver <- rep(0, nrow(cohort_data))
cohort_data$Liver <- ifelse(cohort_data$SUBJECT_ID %in% ld_vector, 1, 0)

# History of Depression
sql_icd9_depression <- sprintf("
    SELECT SUBJECT_ID
    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
    WHERE icd9_code IN ('29021', '29013', '30112', '29621', '29682', '311', '2980', '29633', '29632', '29626', '29625', '29624', '29623', '29620', '29622', '3091', '29634', '29636', '29635', '29631', '29630') ") 
depression_data <- query_exec(sql_icd9_depression, project, max_pages = 1)
depression_vector <- pull(depression_data, SUBJECT_ID) # subject ids of all patients who ever had depression
                
cohort_data$Depression <- rep(0,nrow(cohort_data))
cohort_data$Depression <- ifelse(cohort_data$SUBJECT_ID %in% depression_vector, 1, 0)

# History of Reactive Airway Disease
sql_icd9_reac <- sprintf("
                    SELECT SUBJECT_ID
                    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
                    WHERE icd9_code IN ('49390', '49310', '49320', '49311', '49312', '49321', '49322') ")
reac_data <- query_exec(sql_icd9_reac, project, max_pages = 1)
reac_vector <- pull(reac_data, SUBJECT_ID)

cohort_data$Reactive_airway <- rep(0, nrow(cohort_data))
cohort_data$Reactive_airway <- ifelse(cohort_data$SUBJECT_ID %in% reac_vector, 1, 0)

# History of Cancer
c_icd <- c(seq(1400, 1649, by=1), seq(1700, 1759, by=1), seq(1800, 1809, by = 1), 
                             seq(1820, 1969, by=1), seq(1990, 1999, by=1), seq(2350, 2399, by=1))
c_string_icd <- toString(sprintf("'%s'", c_icd))

sql_icd9_cancer <- sprintf("
SELECT SUBJECT_ID
    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
    WHERE icd9_code IN (%s)", paste(c_string_icd, collapse=", "))
cancer_data <- query_exec(sql_icd9_cancer, project, max_pages = 1)
cancer_vector <- pull(cancer_data, SUBJECT_ID)

cohort_data$Cancer <- rep(0, nrow(cohort_data))
cohort_data$Cancer <- ifelse(cohort_data$SUBJECT_ID %in% cancer_vector, 1, 0)


kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# OPTIMIZE read_in features - <feature> #WILL NEED TO HAVE feature extraction code
will fix to put in extraction code

```{r eval=T, message=F, warning=F}
# read in other OPTIMIZE features and add to cohort

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# EFFECT feature extraction - Depression, Cancer

```{r eval=T, message=F, warning=F}
# Depression

# Cancer

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# EFFECT feature read in - 

```{r eval=T, message=F, warning=F}
# 

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# OPTIMIE features left to get - 

```{r eval=T, message=F, warning=F}
# OPTIMIZE features

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# Imputation of missing values and related plots
add references to dinov's tutorial

```{r eval=T, message=F, warning=F}
library(Amelia)
sim_data.df <- read.csv(file='DATA_MASTER.csv', header=TRUE)
dim(sim_data.df)
amelia.out <- amelia(sim_data.df, m = 5)
amelia.out
amelia.imputed.5 <- amelia.out$imputations[[5]]


imputed_scores <- read.csv(file='DATA_IMPUTE_EFFECT_OPTIME_OPTIMIZE.csv', header=TRUE)
normal_scores <- read.csv(file='DATA_effect_optime_optimize.csv', header=TRUE)
```

# EFFECT Score calculation! and basic summary stats?
```{r eval=T, message=F, warning=F}
# EFFECT
#age, cbv disease, dementia, COPD, BUN + cirrhosis + hemoglobin + resp, SBP, SODIUM
cohort_data$EFFECT <- rep(0, nrow(cohort_data))
for (i in 1:nrow(cohort_data)){
  score <- 0
  #age
  score <- score + cohort_data$AGE[i]
  if( cohort_data$CV_Disease[i] == 1){
    score <- score + 10
  }
  if( cohort_data$Dementia[i] == 1){
    score <- score + 10
  }
  if (cohort_data$COPD[i] == 1){
    score <- score + 20
  }
  # do nothing for hemoglobin i guess
  if(cohort_data$BUN[i] <= 60){
    score <- score + cohort_data$BUN[i]
  }
  if(cohort_data$RESP_RATE[i] >= 20 && cohort_data$RESP_RATE[i] <= 45){
    score <- score + cohort_data$RESP_RATE[i]
  }
  if( cohort_data$cirrhosis[i] == 1){
    score <- score + 25
  }
  # sbp
  if(cohort_data$SBP[i] >= 180){
    score <- score - 60
  } else if(cohort_data$SBP[i] >= 160){
    score <- score - 55
  } else if(cohort_data$SBP[i] >= 140){
    score <- score - 50
  } else if(cohort_data$SBP[i] >= 120){
    score <- score- 45
  } else if(cohort_data$SBP[i] >= 100){
    score <- score - 40
  } else if(cohort_data$SBP[i] >= 90){
    score <- score - 35
  } else if(cohort_data$SBP[i] < 90){
    score <- score - 30
  }
  if(cohort_data$SODIUM[i] < 136){
   score <- score + 10
  }
  cohort_data$EFFECT[i] <- score
}
```

# OPTIME Score
```{r eval=T, message=F, warning=F}
# OPTIMIE
```

# OPTIMIZE Score
```{r eval=T, message=F, warning=F}
# OPTIMIZE
#age, cbv disease, dementia, COPD, BUN + cirrhosis + hemoglobin + resp, SBP, SODIUM
#cohort_data_complete <- read.csv('DATA.csv')

cohort_data$OPTIMIZE <- rep(0, nrow(cohort_data))
for (i in 1:nrow(cohort_data)){
  score <- 0
  #age
  if( cohort_data$AGE[i] < 30){
    score <- score + 0
  } else if( cohort_data$AGE[i] < 40){
    score <- score + 2
  }  else if( cohort_data$AGE[i] < 50){
    score <- score + 5
  }  else if( cohort_data$AGE[i] < 60){
    score <- score + 7
  }  else if( cohort_data$AGE[i] < 70){
    score <- score + 10
  }  else if( cohort_data$AGE[i] < 80){
    score <- score + 12
  }  else if( cohort_data$AGE[i] < 90){
    score <- score + 15 
  } else if( cohort_data$AGE[i] >= 90){
    score <- score + 17 
  }  
  # weight, kg
  if( cohort_data$WEIGHT[i] < 70){
    score <- score + 9
  } else if( cohort_data$WEIGHT[i] < 90){
    score <- score + 7
  } else if( cohort_data$WEIGHT[i] < 110){
    score <- score + 5
  } else if( cohort_data$WEIGHT[i] < 130){
    score <- score + 3
  } else if( cohort_data$WEIGHT[i] >= 130){
    score <- score + 2
  }
  # sbp
  if(cohort_data$SBP[i] < 90){
    score <- score + 24
  } else if(cohort_data$SBP[i] < 110){
    score <- score + 20
  } else if(cohort_data$SBP[i] < 130){
    score <- score + 17
  } else if(cohort_data$SBP[i] < 150){
    score <- score + 13
  } else if(cohort_data$SBP[i] < 170){
    score <- score + 11
  } else if(cohort_data$SBP[i] < 190){
    score <- score + 9
  } else if(cohort_data$SBP[i] < 210){
    score <- score +8
  } else if(cohort_data$SBP[i] < 230){
    score <- score +6
  } else if(cohort_data$SBP[i] < 250){
    score <- score +4
  } else if(cohort_data$SBP[i] < 270){
    score <- score +2
  } else if(cohort_data$SBP[i] >= 270){
    score <- score + 0
  } 
  
  #sodium
  if(cohort_data$SODIUM[i] < 112.5){
   score <- score + 12
  } else if(cohort_data$SODIUM[i] < 117.5){
   score <- score + 10
  } else if(cohort_data$SODIUM[i] < 122.5){
   score <- score + 8
  } else if(cohort_data$SODIUM[i] < 127.5){
   score <- score + 6
  } else if(cohort_data$SODIUM[i] < 132.5){
   score <- score + 4
  } else if(cohort_data$SODIUM[i] < 137.5){
   score <- score + 2
  } else if(cohort_data$SODIUM[i] >= 137.5 ){
   score <- score + 0
  }
  #creatinine 
  if (cohort_data$CREATININE[i] < 0.5){
    score <- score + 0
  } else if (cohort_data$CREATININE[i] < 1.5){
    score <- score + 5
  } else if (cohort_data$CREATININE[i] < 2.5){
    score <- score + 9
  } else if (cohort_data$CREATININE[i] < 3.5){
    score <- score + 14
  } else if (cohort_data$CREATININE[i] >= 3.5){
    score <- score + 19
  }
  
  # history of liver disease 
  if(cohort_data$Liver[i] == 1){
    score <- score + 8
  }
  
  #history of reactive airway disease 
  if( cohort_data$Reactive_airway[i] == 1){
    score <- score + 4
  }
  #history of depression
  if( cohort_data$Depression[i] == 1){
    score <- score + 4
  }
  
  cohort_data$OPTIMIZE[i] <- score
}
```

# Binning patients based on time til death outside of hospital


```{r eval=T, message=F, warning=F}
# Death binning code
# Emily has it and will add if we want it

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# Plots analyzing binned death for scores 
analysis of different scores motives (optimize looking at 6-9 month mortality, ect)


```{r eval=T, message=F, warning=F}
# plots of binned deaths!
```

# Mortality as continuous variable plots and analysis
add references to dinov's tutorial
