---
title: "Acute Heart Failure"
author: "<h3>SOCR MIMIC-III Team</h3>"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
  highlight: tango
number_sections: yes
theme: default
toc: yes
toc_depth: 2
toc_float:
  collapsed: no
smooth_scroll: yes
---
  
*Libraries Used*
```{r message=F, warning=F}
# Data I/O
#library('RPostgreSQL')    # access MIMIC-III
library('reshape2')       # melting dataframes
library('dplyr')          # misc. organization
library('bigrquery')     # Query data from Google BigQuery
# Data preparation
library('psych')          # descriptive stats

# Plots and tables
library('knitr')          # knitting Rmd to HTML; kable() function
library('kableExtra')     # extra formating options for knitr tables
library('ggplot2')        # 2d plotting
library('ggpubr')         # extra formatting options for ggplot
library('gtools')
library("lubridate")

library(survival)

project = 'mimic-iii-244018'
```

# Introduction

Heart failure (HF) is a "multifactorial, systemic disease" in which a network of autoprotective mechanisms activated after cardiac injury cause significant problems with heart function ([Tanai & Frantz, 2015](https://www.ncbi.nlm.nih.gov/pubmed/26673558)). HF is one of the largest contributing factors to mortality in the United States, playing a role in 1 in 9 deaths and accounting for more than $30 billion in annual healthcare expenditures ([CDC.gov](https://www.cdc.gov/dhdsp/data_statistics/fact_sheets/fs_heart_failure.htm), [Mozaffarian et. al](https://www.ncbi.nlm.nih.gov/pubmed/26673558)). In addition to the estimated 5.7 million Americans living with HF, an additional 915,000 are newly diagnosed each year (Mozaffarian et. al). Despite recent advancements in the treatment of underlying risk factors, long-term mortality remains high with less than half of those newly diagnoses patients suriviving for five years (Mozaffarian et. al).

Risk stratification systems have become an important part of patient management, and have the potential to "improve clinical outcome and resource allocation" by "avoiding the overtreatment of low-risk subjects or the early, inappropriate discharge of high-risk patients" ([Passantino et. al, 2015](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4691817/)). In their comprehensive 2015 publication, Andrea Passantino and colleagues review a collection of the "most relevant" risk stratification systems targeted at acute heart failure (AHF) including EFFECT, OPTIMIZE-HF, APACHE-HF, and ESCAPE, among others. Passantino and her team describe the wide array of data sources and techniques used by the original authors to validate these scoring systems, including "public registries, clinical trials, and retrospective data" ([Passantino et. al, 2015](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4691817/)). The heterogeneity of these data sources makes direct performance comparisons difficult.

In this document, we aim to extend the work contributed by Passantino and colleagues and provide a direct comparison of the performance of these AHF risk stratification metrics on a single, unified dataset. To this end, we leverage the Medical Information Mart for Intensive Care ([MIMIC-III](https://mimic.physionet.org/)), a dataset developed by the Massachusetts Institute of Technology Lab for Computational Physiology (MIT-LCP) which contains de-identified health data for more than 40,000 intensive care unit (ICU) patients over the years 2001-2012 ([Johnson et. al, 2012](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4878278/)). It includes much of the same data commonly found in electronic health records (EHR), including demographic information, diagnosis codes, procedures, laboratory values, vital signs, free-text clinical notes, and admission, discharge, and mortality information. Moreover, it is large and comprehensive enough to serve as a proving ground for each of the individual risk stratification systems, allowing a level playing field from which to make direct comparisons of the efficacy and performance of each proposed metric.

In the spirit of [SOCR's](http://www.socr.umich.edu) [open science](https://en.wikipedia.org/wiki/Open_science) initiative, this document and it's parent repository contain the complete end-to-end computational protocol, results, and validation procedures. It is subdivided into the following partitions:

1. *Cohort selection*: criteria for inclusion into the study are codified and explained.
2. *Data extraction*: specific data points are extracted from the larger dataset and transformed for further analysis.
3. *Data manipulation*: extracted data is analyzed and imputed to facilitate scoring.
4. *Scoring*: scoring systems are discussed and applied.
5. *Risk analysis*: the effectiveness of the scoring systems are determined.

# Cohort Selection

As the MIMIC-III dataset contains a wide variety of patients, those suitable for our particular analysis must first be isolated. Based on the manuscripts cited in [Passantino et. al's review](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4691817/), we developed a set of inclusion criteria to select such a cohort. These criteria include:

1. Patients diagnosed with acute heart failure, excluding those with existing chronic heart failure.
2. Patients who survived to ICU discharge, prefering to instead focus on out-of-hospital mortality.
3. Patients who stayed in the ICU between 24 hours and 45 days, yielding enough time to facilitate a number of test results while exluding the few ultra-long term ICU stays present in the dataset.
4. Patients who are between 18 and 89 years of age (inclusive).

These criteria were inspired by the inclusion criteria schema described in the following references:

1. [Auble et. al, A Prediction Rule to Identify Low-risk Patients with Heart Failure. *Academic Emergency Medicine*, 2005.](https://onlinelibrary.wiley.com/doi/pdf/10.1197/j.aem.2004.11.026)  
2. [Abraham et. al, Predictors of In-Hospital Mortality in Patients Hospitalized for Heart Failure: Insights From the Organized Program to Initiate Lifesaving Treatment in Hospitalized Patients With Heart Failure (OPTIMIZE-HF). *Journal of the Americal College of Cardiology*, 2008.](https://www.sciencedirect.com/science/article/pii/S0735109708016720)  
3. [Peterson et. al, A Validated Risk Score for In-Hospital Mortality in Patients With Heart Failure From the American Heart Association Get With the Guidelines Program. *Circulation*, 2009.](https://www.ahajournals.org/doi/pdf/10.1161/CIRCOUTCOMES.109.854877)  
4. [Lee et. al, Prediction of Heart Failure Mortality in Emergent Care: A Cohort Study. *Annals of Internal Medicine*, 2012.](https://annals.org/aim/article-abstract/1170879/prediction-heart-failure-mortality-emergent-care-cohort-study)   
5. [Okazaki et. al, New scoring system (APACHE-HF) for predicting adverse outcomes in patients with acute heart failure: Evaluation of the APACHE II and Modified APACHE II scoring systems. *Journal of Cardiology*, 2014](https://www.sciencedirect.com/science/article/pii/S0914508714000951?via%3Dihub)  
6. [Salah et. al, A novel discharge risk model for patients hospitalised for acute decompensated heart failure incorporating N-terminal pro-B-type natriuretic peptide levels: a European coLlaboration on Acute decompeNsated Heart Failure: ÉLAN-HF Score. *Heart*, 2013](https://heart.bmj.com/content/100/2/115.long)  
7. [Lee et. al, Predicting Mortality Among Patients Hospitalized for Heart Failure. *JAMA*, 2003](https://jamanetwork.com/journals/jama/fullarticle/197670)  
8. [O'Connor, et. al, Triage After Hospitalization With Advanced Heart Failure: The ESCAPE (Evaluation Study of Congestive Heart Failure and Pulmonary Artery Catheterization Effectiveness) Risk Model and Discharge Score, *Jounral of the American College of Cardiology*, 2010](https://www.sciencedirect.com/science/article/pii/S0735109709040595?via%3Dihub)

## Criteria 1: Patients diagnosed with actue heart failure

To select patients with acute heart failure, we first constructed a cohort of all heart failure patients, denoted by the ICD9 code `4280`, then subtracted those that had an additional chronic heart failure diagnosis (codes `42822, 42823, 42832, 42833, 42842 & 42843`).

```{r message=F, warning=F}
sql <- "SELECT *
        FROM [physionet-data.mimiciii_clinical.diagnoses_icd]
        WHERE icd9_code IN ('4280')
        AND hadm_id NOT IN (
            SELECT hadm_id
            FROM [physionet-data.mimiciii_clinical.diagnoses_icd]
            WHERE icd9_code IN ('42822', '42823', '42832', '42833', '42842', '42843') 
            GROUP BY hadm_id
        )"

cohort <- query_exec(sql, project=project)
```

Based on this initial query, there are `r length(unique(cohort$HADM_ID))` unique encounters in our cohort representing `r length(unique(cohort$SUBJECT_ID))` individual subjects.

## Criteria 2: Subject survived to ICU discharge
As our target is predicting out-of-hospital mortality due to acute heart failure, we excude patients who expired in the hospital.
```{r message=F, warning=F}
# Retrieve ADMISSIONS table
adm.tbl <- query_exec(
  sprintf( "SELECT HADM_ID, DISCHTIME, DEATHTIME
        FROM [physionet-data.mimiciii_clinical.admissions] 
        WHERE hadm_id IN (%s)",
      paste(cohort$HADM_ID, collapse=", ")),
  project=project, max_pages = Inf)

# Merge with cohort object
cohort <- merge(cohort, adm.tbl, by="HADM_ID")

# Remove subjects with in hospital date of death
cohort <- cohort[is.na(cohort$DEATHTIME), ]
```

After removing subjects who did not survive to ICU discharge, we are left with `r length(unique(cohort$HADM_ID))` unique encounters representing `r length(unique(cohort$SUBJECT_ID))` individual subjects.

## Criteria 3: Length of stay between 24-hours and 45 days

```{r message=F, warning=F}
# Get admissions table object
admissions.tbl <- query_exec(
  sprintf( "SELECT *
            FROM [physionet-data.mimiciii_clinical.admissions] 
            WHERE hadm_id IN (%s)", 
          paste(cohort$HADM_ID, collapse=", ")),
  project=project, max_pages = Inf)

# Calculate length of stay
admissions.tbl$LOS <- difftime(admissions.tbl$DISCHTIME, admissions.tbl$ADMITTIME, unit='days')

# Merge with cohort object
cohort <- merge(cohort, admissions.tbl[, c('SUBJECT_ID', 'HADM_ID', 'DISCHTIME', 'ADMITTIME', 'LOS')], by=c('SUBJECT_ID', 'HADM_ID'))

# Plot length of stay before removal
fig <- ggplot(cohort, aes(x=LOS)) +
  geom_histogram() + 
  ggtitle('Length of Stay') + xlab('Days')
print(fig)

# Remove encounters where LOS falls outside bounds
cohort <- cohort[cohort$LOS > 1 & cohort$LOS <= 45, ]
```

After removing subjects who did not survive to ICU discharge, we are left with `r length(unique(cohort$HADM_ID))` unique encounters representing `r length(unique(cohort$SUBJECT_ID))` individual subjects.

```{r message=F, warning=F}
sql <- sprintf(
       "SELECT admissions.HADM_ID as admissions_HADM_ID, admissions.ADMITTIME as admissions_ADMITTIME, patients.DOD as DOD, patients.DOD_HOSP as DOD_HOSP, patients.DOD_SSN as DOD_SSN, patients.DOB as patients_DOB
        FROM [physionet-data.mimiciii_clinical.admissions] AS admissions
        JOIN [physionet-data.mimiciii_clinical.patients] AS patients
        ON admissions.subject_id = patients.subject_id
        WHERE admissions.hadm_id IN (%s)",
        paste(cohort$HADM_ID, collapse = ", "))

age.tbl <- query_exec(sql, project=project)
age.tbl$age <- difftime(age.tbl$admissions_ADMITTIME, age.tbl$patients_DOB, units='days')

age.tbl <- data.frame(age.tbl[, c('admissions_HADM_ID', 'age', 'DOD', 'DOD_HOSP', 'DOD_SSN')])
colnames(age.tbl) <- c('HADM_ID', 'AGE', 'DOD', 'DOD_HOSP', 'DOD_SSN')
cohort <- merge(cohort, age.tbl, by="HADM_ID")
cohort$AGE <- as.numeric(cohort$AGE) / 365

cohort <- cohort[cohort$AGE < 90 & cohort$AGE >= 18, ]

# Plot length of stay before removal
fig <- ggplot(cohort, aes(x=AGE)) +
  geom_histogram() + 
  ggtitle('Subject Age') + xlab('Years')
print(fig)
```

The final cohort consists of `r nrow(cohort)` encounters.

# Data extraction

```{r message=F, warning=F}
pull.last.event <- function(hadm_id, itemcodes, return.fields, table, project, max_pages=Inf) {
  sql <- sprintf("
                 SELECT *
                 FROM [physionet-data.mimiciii_clinical.%s] AS table1
                 INNER JOIN (
                   SELECT hadm_id, MAX(charttime), MAX(row_id)
                   FROM [physionet-data.mimiciii_clinical.%s]
                   WHERE itemid IN (%s)
                   AND hadm_id IN (%s)
                   GROUP BY hadm_id
                 ) AS table2
                 ON table1.row_id = table2.f1_",
                 table, table, 
                 paste(itemcodes, collapse=", "),
                 paste(hadm_id, collapse=", ")
  )
  
  data <- query_exec(sql, project=project, max_pages=max_pages)
  colnames(data) <- gsub('table[0-9]_', '', colnames(data))
  return(data[ , return.fields])
}
```

# Discussion of EFFECT, OPTIME, OPTIMIZE

## EFFECT Scoring System
The Enhanced Feedback for Effective Cardiac Treatment (EFFECT) scoring system is a derivation and validation of a clinical model that aims to predict mortality among patients hospitalized for heart failure. The objective is to identify high-risk heart failure patients at hospital discharge, thus allowing more effective triage to management strategies. In order to apply the EFFECT scoring system on the MIMIC-III dataset, we retrieved the cohort patients' age, systolic blood pressure, elevated blood urea nitrogen level, sodium concentration, cerebrovascular disease, hepatic cirrhosis, history of cancer and hemoglobin level. {r} [(M. O'Connor 2010)] {https://www.sciencedirect.com/science/article/pii/S0735109709040595}

## OPTIME-CHF Scoring System
The Outcomes of a Prospective Trial of Intravenous Milrinone for Exacerbations of Chronic Heart Failure (OPTIME-CHF) trial aims to evaluate the treatment strategies in the care of population with acute exacerbations of chronic heart failure. The primary goal of this scoring system is a reduction in the total hospital days for cardiovascular events within 60 days after therapy. In order to apply the OPTIME-CHF scoring system on the MIMIC-III dataset, we retrieved the cohort patients' age, elevated blood urea nitrogen level and systolic blood pressure. {r} [(Duke Clinical Research Institue 2004)] {https://www.ncbi.nlm.nih.gov/pubmed/15599835}

Note: The OPTIME-CHF scoring system also includes NYHA Class as one of the criteria, but since it is not included as part of the MIMIC-III dataset and is a classification system that is highly subjective and varies strongly depending on the doctor diagnosing, it is left out of our calculations in this study.

## OPTIMIZE-HF Scoring System
The Organized Program to Initiate Lifesaving Treatment in Hospitalized Patients with Heart Failure (OPTIMIZE-HF) is a comprehensive hospital-based registry and performance-improvement program designed to predict in-hospital mortality in patients hospitalized for heart failure. In order to apply the OPTIMIZE-HF scoring system on the MIMIC-III dataset, we retrieved the cohort patients' age, weight, systolic blood pressure, creatinine level, liver disease, history of depression and reactive airway disease. {r} [(Duke Clinical Research Institue 2008)] {https://www.ncbi.nlm.nih.gov/pubmed/18926148}

## Extracting itemcodes

### BUN item codes
```{r}
sql = "SELECT ITEMID FROM [physionet-data.mimiciii_clinical.d_items] 
        WHERE LABEL LIKE \"%BUN%\""
bun.itemcodes = query_exec(sql, project=project)
bun.itemcodes = bun.itemcodes$ITEMID
```
 
### Creatinine item code
```{r}
sql = "SELECT ITEMID FROM [physionet-data.mimiciii_clinical.d_items] 
        WHERE LABEL LIKE \"%creatinine%\" OR LABEL LIKE \"%Creatinine%\" "
creatinine.itemcodes = query_exec(sql, project=project)
creatinine.itemcodes = creatinine.itemcodes$ITEMID
```

### Hemoglobin item codes
```{r}
sql = "SELECT ITEMID FROM [physionet-data.mimiciii_clinical.d_items] 
        WHERE LABEL LIKE \"%hemoglobin%\" OR LABEL LIKE \"%Hemoglobin%\" "
hemoglobin.itemcodes = query_exec(sql, project=project)
hemoglobin.itemcodes = hemoglobin.itemcodes$ITEMID
```

### Sodium itemcodes
```{r}
sql = "SELECT ITEMID FROM [physionet-data.mimiciii_clinical.d_items] 
        WHERE LABEL LIKE \"%sodium%\" OR LABEL LIKE \"%Sodium%\" "
sodium.itemcodes = query_exec(sql, project=project)
sodium.itemcodes = sodium.itemcodes$ITEMID
```

## Extract lab results

This pull.last.event function is going to be used when extracting lab results
```{r message=F, warning=F}
pull.last.event <- function(hadm_id, itemcodes, return.fields, table, project, max_pages=Inf) {
  sql <- sprintf("
                 SELECT *
                 FROM [physionet-data.mimiciii_clinical.%s] AS table1
                 INNER JOIN (
                   SELECT hadm_id, MAX(charttime), MAX(row_id)
                   FROM [physionet-data.mimiciii_clinical.%s]
                   WHERE itemid IN (%s)
                   AND hadm_id IN (%s)
                   GROUP BY hadm_id
                 ) AS table2
                 ON table1.row_id = table2.f1_",
                 table, table, 
                 paste(itemcodes, collapse=", "),
                 paste(hadm_id, collapse=", ")
  )
  
  data <- query_exec(sql, project=project, max_pages=max_pages)
  colnames(data) <- gsub('table[0-9]_', '', colnames(data))
  return(data[ , return.fields])
}
```

### Systolic Blood Pressure
```{r message=F, warning=F}
sbp.itemcodes <- c( 6, 51, 442, 3313, 3315, 3317, 3321, 3323, 3325, 6701, 228152, 224167, 227243, 220050, 220179, 225309 )
return.fields <- c('HADM_ID', 'VALUENUM')
data.sbp <- pull.last.event(cohort$HADM_ID, sbp.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.sbp)[2] = "SBP"
```

### Blood Urea Nitrogen
```{r}
return.fields <- c('HADM_ID', 'VALUENUM')
data.bun <- pull.last.event(cohort$HADM_ID, bun.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.bun)[2] = "BUN"
```

### Creatinine
```{r}
return.fields <- c('HADM_ID', 'VALUENUM')
data.creatinine <- pull.last.event(cohort$HADM_ID, creatinine.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.creatinine)[2] = "Creatinine"
```

### Weight
```{r}
weight.itemcodes <- c(226531, 763)
return.fields <- c('HADM_ID', 'VALUENUM')
data.weight <- pull.last.event(cohort$HADM_ID, weight.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.weight)[2] = "Weight"
```

### Sodium
```{r}
sodium.itemcodes <- c(220645)
return.fields <- c('HADM_ID', 'VALUENUM')
data.sodium <- pull.last.event(cohort$HADM_ID, sodium.itemcodes, return.fields, 'chartevents', project, max_pages=1)

colnames(data.sodium)[2] = "Sodium"
```

### Hepatic Cirrhosis
[5715 is the icd9 code for hepatic cirrhosis](http://www.icd9data.com/2012/Volume1/520-579/570-579/571/571.5.htm)
```{r}
sql <- "SELECT hadm_id
        FROM [physionet-data.mimiciii_clinical.diagnoses_icd]
        WHERE icd9_code = \"5715\" "

cirr.cohort <- query_exec(sql, project=project)

```

## Combine the data
```{r}
cohort <- left_join(cohort, data.sbp, by='HADM_ID') %>%
                left_join(., data.creatinine, by='HADM_ID') %>%
                left_join(., data.weight, by='HADM_ID') %>%
                left_join(., data.sodium, by='HADM_ID')

cohort$cirr = FALSE
cohort$cirr = cohort$HADM_ID %in% cirr.cohort$hadm_id
```

### Liver Disease
```{r eval=T, message=F, warning=F}
# History of Liver Disease
sql_icd9_liver <- sprintf("
                    SELECT SUBJECT_ID
                    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
                    WHERE icd9_code IN ('570', '5710', '5712', '5713', '5715', '5728', '5807', '5730', '5738', '5739', '5718', '5719') ")
ld_data <- query_exec(sql_icd9_liver, project, max_pages = 1)
ld_vector <- pull(ld_data, SUBJECT_ID) # subject ids for history of liver disease

cohort$Liver <- rep(0, nrow(cohort))
cohort$Liver <- ifelse(cohort$SUBJECT_ID %in% ld_vector, 1, 0)
```

### History of Depression
We referenced [Appendix A.2 from Healthcare Cost and Utilization Project (HCUP) Statistical Briefs](https://www.ncbi.nlm.nih.gov/books/NBK409512/table/sb216.t6/) in selection of ICD9 codes related to depression. 
```{r eval=T, message=F, warning=F}
sql_icd9_depression <- sprintf("
    SELECT SUBJECT_ID
    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
    WHERE icd9_code IN ('29021', '29013', '30112', '29621', '29682', '311', '2980', '29633', '29632', '29626', '29625', '29624', '29623', '29620', '29622', '3091', '29634', '29636', '29635', '29631', '29630') ") 

depression_data <- query_exec(sql_icd9_depression, project, max_pages = 1)
depression_vector <- pull(depression_data, SUBJECT_ID) # subject ids of all patients who ever had depression
                
cohort$Depression <- rep(0,nrow(cohort))
cohort$Depression <- ifelse(cohort$SUBJECT_ID %in% depression_vector, 1, 0)
```

### History of Reactive Airway Disease
```{r eval=T, message=F, warning=F}
sql_icd9_reac <- sprintf("
                    SELECT SUBJECT_ID
                    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
                    WHERE icd9_code IN ('49390', '49310', '49320', '49311', '49312', '49321', '49322') ")
reac_data <- query_exec(sql_icd9_reac, project, max_pages = 1)
reac_vector <- pull(reac_data, SUBJECT_ID)

cohort$Reactive_airway <- rep(0, nrow(cohort))
cohort$Reactive_airway <- ifelse(cohort$SUBJECT_ID %in% reac_vector, 1, 0)
```

### History of Cancer
We referenced [Weiner, M.D. et. al’s paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1480106/?fbclid=IwAR2rOQKA9k-61IqwWto3c_6mOcNfdBUzbd2R6y0h6JhCoJTaJ3LynO9B2Fw) for which ICD9 codes to extract for a subject’s history of cancer. We used all groups of cancer so to include anyone with a history of cancer. 
```{r eval=T, message=F, warning=F}
c_icd <- c(seq(1400, 1649, by=1), seq(1700, 1759, by=1), seq(1800, 1809, by = 1), 
                             seq(1820, 1969, by=1), seq(1990, 1999, by=1), seq(2350, 2399, by=1))
c_string_icd <- toString(sprintf("'%s'", c_icd))

sql_icd9_cancer <- sprintf("
SELECT SUBJECT_ID
    FROM [physionet-data.mimiciii_clinical.diagnoses_icd] 
    WHERE icd9_code IN (%s)", paste(c_string_icd, collapse=", "))
cancer_data <- query_exec(sql_icd9_cancer, project, max_pages = 1)
cancer_vector <- pull(cancer_data, SUBJECT_ID)

cohort$Cancer <- rep(0, nrow(cohort))
cohort$Cancer <- ifelse(cohort$SUBJECT_ID %in% cancer_vector, 1, 0)


kable(head(cohort), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

# Imputation of missing values and related plots
add references to dinov's tutorial

```{r eval=T, message=F, warning=F}
library(Amelia)
sim_data.df <- read.csv(file='DATA_MASTER.csv', header=TRUE)
dim(sim_data.df)
amelia.out <- amelia(sim_data.df, m = 5)
amelia.out
amelia.imputed.5 <- amelia.out$imputations[[5]]

```

# EFFECT Score Calculation
```{r eval=T, message=F, warning=F}
# EFFECT
#age, cbv disease, dementia, COPD, BUN + cirrhosis + hemoglobin + resp, SBP, SODIUM
cohort$EFFECT <- rep(0, nrow(cohort))
for (i in 1:nrow(cohort)){
  score <- 0
  #age
  score <- score + cohort$AGE[i]
  if( cohort$CV_Disease[i] == 1){
    score <- score + 10
  }
  if( cohort$Dementia[i] == 1){
    score <- score + 10
  }
  if (cohort$COPD[i] == 1){
    score <- score + 20
  }
  # do nothing for hemoglobin i guess
  if(cohort$BUN[i] <= 60){
    score <- score + cohort$BUN[i]
  }
  if(cohort$RESP_RATE[i] >= 20 && cohort$RESP_RATE[i] <= 45){
    score <- score + cohort$RESP_RATE[i]
  }
  if( cohort$cirrhosis[i] == 1){
    score <- score + 25
  }
  # sbp
  if(cohort$SBP[i] >= 180){
    score <- score - 60
  } else if(cohort$SBP[i] >= 160){
    score <- score - 55
  } else if(cohort$SBP[i] >= 140){
    score <- score - 50
  } else if(cohort$SBP[i] >= 120){
    score <- score- 45
  } else if(cohort$SBP[i] >= 100){
    score <- score - 40
  } else if(cohort$SBP[i] >= 90){
    score <- score - 35
  } else if(cohort$SBP[i] < 90){
    score <- score - 30
  }
  if(cohort$SODIUM[i] < 136){
   score <- score + 10
  }
  cohort$EFFECT[i] <- score
}
```

# OPTIME Score Calculation
```{r eval=T, message=F, warning=F}
# OPTIME
cohort$OPTIME <- rep(0, nrow(cohort))
for (i in 1:nrow(cohort)){
  score = 0
  if(is.na(AGE)){
  }else if(AGE > 20 && AGE <= 30){
    score = score + 8
  }else if(AGE > 30 && AGE <= 40){
    score = score + 17
  }else if(AGE > 40 && AGE <= 50){
    score = score + 25
  }else if(AGE > 50 && AGE <= 60){
    score = score + 33
  }else if(AGE > 60 && AGE <= 70){
    score = score + 41
  }else if(AGE > 70 && AGE <= 80){
    score = score + 50
  }else if(AGE > 80 && AGE <= 90){
    score = score + 58
  }
  
  if(is.na(sodium)){
  }else if(sodium < 115){
    score = score + 79
  }else if(sodium < 120){
    score = score + 69
  }else if(sodium < 125){
    score = score + 59
  }else if(sodium < 130){
    score = score + 49
  }else if(sodium < 135){
    score = score + 30
  }else if(sodium < 140){
    score = score + 20
  }else if(sodium < 145){
    score = score + 10
  }
  
  if(is.na(SBP)){
  }else if(SBP < 80){
    score = score + 94
  }else if(SBP < 90){
    score = score + 86
  }else if(SBP < 100){
    score = score + 77
  }else if(SBP < 110){
    score = score + 69
  }else if(SBP < 120){
    score = score + 60
  }else if(SBP < 130){
    score = score + 51
  }else if(SBP < 140){
    score = score + 43
  }else if(SBP < 150){
    score = score + 34
  }else if(SBP < 160){
    score = score + 26
  }else if(SBP < 170){
    score = score + 17
  }else if(SBP < 180){
    score = score + 9
  }
  
  if(is.na(BUN)){
  }else if(BUN < 5){
    score = score + 10
  }else if(BUN < 10){
    score = score + 20
  }else if(BUN < 15){
    score = score + 30
  }else if(BUN < 20){
    score = score + 40
  }else if(BUN < 25){
    score = score + 50
  }else if(BUN < 30){
    score = score + 60
  }else if(BUN < 35){
    score = score + 70
  }else if(BUN < 40){
    score = score + 80
  }else if(BUN < 45){
    score = score + 90
  }else{
    score = score + 100
  }
  
  cohort$OPTIME[i] <- score
}


```

# OPTIMIZE Score Calculation
```{r eval=T, message=F, warning=F}
# OPTIMIZE
#age, cbv disease, dementia, COPD, BUN + cirrhosis + hemoglobin + resp, SBP, SODIUM
#cohort_complete <- read.csv('DATA.csv')

cohort$OPTIMIZE <- rep(0, nrow(cohort))
for (i in 1:nrow(cohort)){
  score <- 0
  #age
  if( cohort$AGE[i] < 30){
    score <- score + 0
  } else if( cohort$AGE[i] < 40){
    score <- score + 2
  }  else if( cohort$AGE[i] < 50){
    score <- score + 5
  }  else if( cohort$AGE[i] < 60){
    score <- score + 7
  }  else if( cohort$AGE[i] < 70){
    score <- score + 10
  }  else if( cohort$AGE[i] < 80){
    score <- score + 12
  }  else if( cohort$AGE[i] < 90){
    score <- score + 15 
  } else if( cohort$AGE[i] >= 90){
    score <- score + 17 
  }  
  # weight, kg
  if( cohort$WEIGHT[i] < 70){
    score <- score + 9
  } else if( cohort$WEIGHT[i] < 90){
    score <- score + 7
  } else if( cohort$WEIGHT[i] < 110){
    score <- score + 5
  } else if( cohort$WEIGHT[i] < 130){
    score <- score + 3
  } else if( cohort$WEIGHT[i] >= 130){
    score <- score + 2
  }
  # sbp
  if(cohort$SBP[i] < 90){
    score <- score + 24
  } else if(cohort$SBP[i] < 110){
    score <- score + 20
  } else if(cohort$SBP[i] < 130){
    score <- score + 17
  } else if(cohort$SBP[i] < 150){
    score <- score + 13
  } else if(cohort$SBP[i] < 170){
    score <- score + 11
  } else if(cohort$SBP[i] < 190){
    score <- score + 9
  } else if(cohort$SBP[i] < 210){
    score <- score +8
  } else if(cohort$SBP[i] < 230){
    score <- score +6
  } else if(cohort$SBP[i] < 250){
    score <- score +4
  } else if(cohort$SBP[i] < 270){
    score <- score +2
  } else if(cohort$SBP[i] >= 270){
    score <- score + 0
  } 
  
  #sodium
  if(cohort$SODIUM[i] < 112.5){
   score <- score + 12
  } else if(cohort$SODIUM[i] < 117.5){
   score <- score + 10
  } else if(cohort$SODIUM[i] < 122.5){
   score <- score + 8
  } else if(cohort$SODIUM[i] < 127.5){
   score <- score + 6
  } else if(cohort$SODIUM[i] < 132.5){
   score <- score + 4
  } else if(cohort$SODIUM[i] < 137.5){
   score <- score + 2
  } else if(cohort$SODIUM[i] >= 137.5 ){
   score <- score + 0
  }
  #creatinine 
  if (cohort$CREATININE[i] < 0.5){
    score <- score + 0
  } else if (cohort$CREATININE[i] < 1.5){
    score <- score + 5
  } else if (cohort$CREATININE[i] < 2.5){
    score <- score + 9
  } else if (cohort$CREATININE[i] < 3.5){
    score <- score + 14
  } else if (cohort$CREATININE[i] >= 3.5){
    score <- score + 19
  }
  
  # history of liver disease 
  if(cohort$Liver[i] == 1){
    score <- score + 8
  }
  
  #history of reactive airway disease 
  if( cohort$Reactive_airway[i] == 1){
    score <- score + 4
  }
  #history of depression
  if( cohort$Depression[i] == 1){
    score <- score + 4
  }
  
  cohort$OPTIMIZE[i] <- score
}
```

# Pulling mortality after imputing rest of the data 
```{r eval=T, message=F, warning=F}
# Calculate days between date of discharge and date of death
cohort_data$DISCH_to_DOD = difftime(cohort_data$DOD, cohort_data$DISCHTIME, units='days')
# round up to nearest integer to represent day
ceiling(cohort_data$DISCH_to_DOD)
```

# Binning patients based on time until death outside of hospital
```{r eval=T, message=F, warning=F}
# Expired within 2 weeks of hospital discharge
cohort_data$DOD_BIN <- ifelse(cohort_data$DISCH_to_DOD < 14, "14days", "") 
# Expired between 2 weeks and 2 months
cohort_data$DOD_BIN <- ifelse(cohort_data$DISCH_to_DOD > 13 & cohort_data$DISCH_to_DOD < 61, "60days", cohort_data$DOD_BIN)
# Expired between 2 months and 1 year
cohort_data$DOD_BIN <- ifelse(cohort_data$DISCH_to_DOD > 60 & cohort_data$DISCH_to_DOD < 366, "365days", cohort_data$DOD_BIN)
# Expired between 1 year and 2 years
cohort_data$DOD_BIN <- ifelse(cohort_data$DISCH_to_DOD > 365 & cohort_data$DISCH_to_DOD < 731, "730days", cohort_data$DOD_BIN)
# Expired after 2 years
cohort_data$DOD_BIN <- ifelse(cohort_data$DISCH_to_DOD > 730, "after_2years", cohort_data$DOD_BIN)
# No DOD (presumed alive)
cohort_data$DOD_BIN <- ifelse(is.na(cohort_data$DISCH_to_DOD), "no_dod", cohort_data$DOD_BIN)
```

# Analyzing mean EFFECT scores by bin

```{r eval=T, message=F, warning=F}
effect_mean <- aggregate(cohort_data$EFFECT, by=list(cohort_data$DOD_BIN), FUN=mean)
names(effect_mean) <- c("DOD_BIN", "mean")
effect_se <- ddply(data, c("DOD_BIN"), summarise,
               N    = length(EFFECT),
               mean = mean(EFFECT),
               sd   = sd(EFFECT),
               se   = sd / sqrt(N)
)

ggplot(effect_se, aes(x=DOD_BIN, y=mean)) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.1) +
    geom_line() +
    geom_point()
```

# Analyzing mean OPTIME scores by bin

```{r eval=T, message=F, warning=F}
optime_mean <- aggregate(data$OPTIME, by=list(data$DOD_BIN), FUN=mean)
names(optime_mean) <- c("DOD_BIN", "mean")
optime_se <- ddply(data, c("DOD_BIN"), summarise,
               N    = length(OPTIME),
               mean = mean(OPTIME),
               sd   = sd(OPTIME),
               se   = sd / sqrt(N)
)

ggplot(optime_se, aes(x=DOD_BIN, y=mean)) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.1) +
    geom_line() +
    geom_point()
```

# Analyzing mean OPTIMIZE scores by bin

```{r eval=T, message=F, warning=F}
optimize_mean <- aggregate(data$OPTIMIZE, by=list(data$DOD_BIN), FUN=mean)
names(optimize_mean) <- c("DOD_BIN", "mean")
optimize_se <- ddply(data, c("DOD_BIN"), summarise,
               N    = length(OPTIMIZE),
               mean = mean(OPTIMIZE),
               sd   = sd(OPTIMIZE),
               se   = sd / sqrt(N)
)

ggplot(optimize_se, aes(x=DOD_BIN, y=mean)) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.1) +
    geom_line() +
    geom_point()
```

# Mortality as continuous variable plots and survival analysis
The survival analysis code was referenced from the Scientific Methods for Health Sciences book: [SMHS SurvivalAnalysis](http://wiki.socr.umich.edu/index.php/SMHS_SurvivalAnalysis)

## EFFECT
```{r eval=T, message=F, warning=F}
EFFECT = cohort$mortality
length(EFFECT)

# Below is the process of survival analysis

# We need to determine the mortatility status first
# status 0 = alive, 1 = dead in hospital
cohort$status <- rep(x=0, times = nrow(cohort))
cohort$status <- ifelse(cohort$mortality>0,yes=cohort$status, no=cohort$status+1)
pd_surv_fit <- survfit(Surv(EFFECT,status)~1, data=cohort)
plot(pd_surv_fit, xlab="EFFECT-HF score", ylab="Survival Probability")

summary(pd_surv_fit)
```

## OPTIME
```{r eval=T, message=F, warning=F}
OPTIME = cohort$mortality
length(OPTIME)

# Below is the process of survival analysis

# We need to determine the mortatility status first
# status 0 = alive, 1 = dead in hospital
cohort$status <- rep(x=0, times = nrow(cohort))
cohort$status <- ifelse(cohort$mortality>0,yes=cohort$status, no=cohort$status+1)
pd_surv_fit <- survfit(Surv(OPTIME,status)~1, data=cohort)
plot(pd_surv_fit, xlab="OPTIME-CHF score", ylab="Survival Probability")

summary(pd_surv_fit)
```

## OPTIMIZE
```{r eval=T, message=F, warning=F}
OPTIMIZE = cohort$mortality
length(OPTIMIZE)

# Below is the process of survival analysis

# We need to determine the mortatility status first
# status 0 = alive, 1 = dead in hospital
cohort$status <- rep(x=0, times = nrow(cohort))
cohort$status <- ifelse(cohort$mortality>0,yes=cohort$status, no=cohort$status+1)
pd_surv_fit <- survfit(Surv(OPTIMIZE,status)~1, data=cohort)
plot(pd_surv_fit, xlab="OPTIMIZE-HF score", ylab="Survival Probability")

summary(pd_surv_fit)
```
#Data Extraction
##Packages and Loading Dataset
Libraries used:
```{r message=F, warning=F}
#Data extraction
require(bigrquery)
# Data I/O
require('RPostgreSQL')    # access MIMIC-III
require('reshape2')       # melting dataframes
require('dplyr')          # misc. organization
# Data imputation
require('mi')
# Plots and tables
require('knitr')          # knitting Rmd to HTML; kable() function
require('kableExtra')     # extra formating options for knitr tables
require('ggplot2')        # 2d plotting
#Survival analysis
require(survival)
require(survminer)
```
Loading raw data:
```{r eval=T, message=F, warning=F}
#cohort <- read.csv('MIMIC_FinalCohort.csv')
cohort <- read.csv('newcohort_with_dod.csv')
kable(head(cohort), caption="Sample of raw cohort data") %>%
  kable_styling(bootstrap_options='striped')
cohort_apache <- read.csv('MIMIC_APACHE_data.csv')
kable(head(cohort_apache), caption="Sample of raw APACHE data") %>%
  kable_styling(bootstrap_options='striped')
```
##Data Pulling
The function used to pull data based on a given set of item IDs is:

```{r message=F, warning=F}
pull.last.event <- function(hadm_id, itemcodes, return.fields, table, project, max_pages=Inf) {
  sql <- sprintf("
                 SELECT *
                 FROM [physionet-data.mimiciii_clinical.%s] AS table1
                 INNER JOIN (
                 SELECT hadm_id, MAX(charttime), MAX(row_id)
                 FROM [physionet-data.mimiciii_clinical.%s]
                 WHERE itemid IN (%s)
                 AND hadm_id IN (%s)
                 GROUP BY hadm_id
                 ) AS table2
                 ON table1.row_id = table2.f1_",
                 table, table, 
                 paste(itemcodes, collapse=", "),
                 paste(hadm_id, collapse=", ")
  )
  
  data <- query_exec(sql, project=project, max_pages=max_pages)
  colnames(data) <- gsub('table[0-9]_', '', colnames(data))
  return(data[ , return.fields])
}
```

This function was copied from previous work available on the MIMIC-III SOCR github.

The process for determining what item IDs to use was primarily based on the information available from the d_items table within MIMIC, which lists a description of each item ID included in the data set.  We searched through the available item IDs to find those most similar to the selected features. The code used to query the d_items table was:
```{r message=F, warning=F}
#Finding itemid (left commented)
#project <- project # change this to your project
#sql <- "SELECT * FROM [physionet-data.mimiciii_clinical.d_items] WHERE LABEL LIKE '%Hematocrit%'"    #<---- change the variable name if you want find other related itemid
#data <- query_exec(sql, project=project)
#kable(data) %>% kable_styling(bootstrap_options = 'striped')
```
The item IDs used for each feature were:

age: Age data was already included as part of the cohort (for details see the cohort selection document on the MIMIC-III github)

heart rate: (211, 220045)

creatinine: (50912)

hematocrit: (50810, 51221)

Glasgo comma score: (198)

potassium: (227464,44711, 46223, 1535, 41956, 220640, 226535, 3792, 829, 227442)

sodium: (220645, 1536, 3803, 837, 228389, 226534, 228390)

systolic blood pressure: (6, 51, 442, 445, 480, 482, 484, 492, 3313, 3315, 3317, 3321, 3323, 3325, 6701, 228152, 224167, 227243, 220050, 220179, 225309, 220059, 226850, 226852, 220179)

diastolic blood pressure: (8364,8368,8440,8503,8504,8505,8506,8441,8444,8445,8446,8448,8502,8507,8508,8555,228151,220060,226851,226853,227242,224643,220051,220180,225310)

For certain features, most notably SBP and DBP, we found a large degree of missingness in the data, and so included more item IDs to try to minimize missingness. 

The code used to pull and combine the data data was:
```{r}
#####Data for other scoring systems: cv disease, COPD, dementia
#pulling data for cerebrovascular disease:

sql <- "SELECT *
    FROM [physionet-data.mimiciii_clinical.diagnoses_icd]
    WHERE icd9_code IN ('430', '431', '4320', '4321', '4329', '43300', '43301', '43310', '43311', '43320', '43321', '43330', '43331', '43380', '43381', '43390', '43391', '43400', '43401', '43410', '43411', '43490', '43491', '4350', '4351', '4352', '4353', '4358', '4359', '436', '4370', '4371', '4372', '4373', '4374', '4375', '4376', '4377', '4378', '4379', '4380', '43810', '43811', '43812', '43813', '43814', '43819', '43820', '43821', '43822', '43830', '43831', '43832', '43840', '43841', '43842', '43850', '43851', '43852', '43853', '4386', '4387', '43881', '43882', '43883', '43884', '43885', '43889', '4389', '436', '4371', '4378', '4379', '4380', '4386', '4387', '4389', '43810', '43811', '43812', '43813', '43814', '43819', '43820', '43821', '43822', '43830', '43831', '43832', '43840', '43841', '43842', '43850', '43851', '43852', '43853', '43881', '43882', '43883', '43884', '43885', '43889', '67400', '67401', '67402', '67403', '67404', '74781', '99702')"
cerebrovas <- query_exec(sql, project=project)
cohort_cerebrovas <- merge(cohort, cerebrovas, by='HADM_ID')
cohort$CV_Disease <- ifelse(cohort$HADM_ID %in% cohort_cerebrovas$HADM_ID, 1, 0)

#pulling data for dementia:

sql <- "SELECT *
    FROM [physionet-data.mimiciii_clinical.diagnoses_icd]
WHERE icd9_code IN ('2900', '29010', '29011', '29012', '29013', '29020', '29021', '2903', '29040', '29041', '29042', '29043', '2908', '2909', '2903', '2912', '29010', '29011', '29012', '29013', '29020', '29021', '29040', '29041', '29042', '29043', '29282', '29410', '29411', '29420', '29421', '33119', '33182')"
dementia <- query_exec(sql, project=project)
cohort_dementia <- merge(cohort, dementia, by='HADM_ID')
cohort$Dementia <- ifelse(cohort$HADM_ID %in% cohort_dementia$HADM_ID, 1, 0)

#pulling data for copd:

sql <- "SELECT *
    FROM [physionet-data.mimiciii_clinical.diagnoses_icd]
WHERE icd9_code IN ('490', '4910', '4911', '49120', '49121', '49122', '4918', '4919', '4920', '4928', '49300', '49301', '49301', '49310', '49311', '49312', '49320', '49321', '49322', '49381', '49382', '49390', '49391', '49392', '4940', '4941', '4950', '4951', '4952', '4953', '4954', '4955', '4956', '4957', '4958', '4959', '496')"
COPD <- query_exec(sql, project=project)
cohort_COPD <- merge(cohort, COPD, by='HADM_ID')
cohort$COPD <- ifelse(cohort$HADM_ID %in% cohort_COPD$HADM_ID, 1, 0)

#####APACHE specific data

#pull heart rate data
hr.itemcodes <- c(211, 220045)
return.fields <- c('SUBJECT_ID', 'HADM_ID', 'ITEMID', 'CHARTTIME', 'VALUENUM', 'VALUEUOM')
data_hr <- pull.last.event(cohort$HADM_ID, hr.itemcodes, return.fields, 'chartevents', project, max_pages=1)

#pull creatinine
creatinine.itemcodes <- c(50912)
return.fields <- c('SUBJECT_ID', 'HADM_ID', 'ITEMID', 'CHARTTIME', 'VALUENUM', 'VALUEUOM')
data_creatinine <- pull.last.event(cohort$HADM_ID, creatinine.itemcodes, return.fields, 'labevents', project, max_pages=1)

#pull hematocrit data
hematocrit.itemcodes <- c(50810, 51221)
return.fields <- c('SUBJECT_ID', 'HADM_ID', 'ITEMID', 'CHARTTIME', 'VALUENUM', 'VALUEUOM')
data_hematocrit <- pull.last.event(cohort$HADM_ID, hematocrit.itemcodes, return.fields, 'labevents', project, max_pages=1)

# pull GCS data
gcs.itemcodes <- c(198)
return.fields <- c('SUBJECT_ID', 'HADM_ID', 'ITEMID', 'CHARTTIME', 'VALUENUM', 'VALUEUOM')
data_GCS <- pull.last.event(cohort$HADM_ID, gcs.itemcodes, return.fields, 'chartevents', project, max_pages=1)

# pull potassium data
potassium.itemcodes <- c(227464,44711, 46223, 1535, 41956, 220640, 226535, 3792, 829, 227442)
return.fields <- c('SUBJECT_ID', 'HADM_ID', 'ITEMID', 'CHARTTIME', 'VALUENUM', 'VALUEUOM')
data_potassium <- pull.last.event(cohort$HADM_ID, potassium.itemcodes, return.fields, 'chartevents', project, max_pages=1)

#pull sodium data
sodium.itemcodes <- c(220645, 1536, 3803, 837, 228389, 226534, 228390)
return.fields <- c('SUBJECT_ID', 'HADM_ID', 'ITEMID', 'CHARTTIME', 'VALUENUM', 'VALUEUOM')
data_sodium <- pull.last.event(cohort$HADM_ID, sodium.itemcodes, return.fields, 'chartevents', project, max_pages=1)


#pull dbp data
dbp.itemcodes <- c(8364,8368,8440,8503,8504,8505,8506,8441,8444,8445,8446,8448,8502,8507,8508,8555,228151,220060,226851,226853,227242,224643,220051,220180,225310)
return.fields <- c('SUBJECT_ID', 'HADM_ID', 'ITEMID', 'CHARTTIME', 'VALUENUM', 'VALUEUOM')
data_dbp <- pull.last.event(cohort$HADM_ID, dbp.itemcodes, return.fields, 'chartevents', project, max_pages=1)

#pull sbp data
sbp.itemcodes <- c(6, 51, 442, 445, 480, 482, 484, 492, 3313, 3315, 3317, 3321, 3323, 3325, 6701, 228152, 224167, 227243, 220050, 220179, 225309, 220059, 226850, 226852, 220179)
return.fields <- c('SUBJECT_ID', 'HADM_ID', 'ITEMID', 'CHARTTIME', 'VALUENUM', 'VALUEUOM')
data_sbp <- pull.last.event(cohort$HADM_ID, sbp.itemcodes, return.fields, 'chartevents', project, max_pages=1)


#join together new data
data_hr$HR <- data_hr$VALUENUM
data_creatinine$creatinine <- data_creatinine$VALUENUM
data_hematocrit$hematocrit <- data_hematocrit$VALUENUM
data_GCS$gcs <- data_GCS$VALUENUM
data_potassium$potassium <- data_potassium$VALUENUM
data_sodium$sodium <- data_sodium$VALUENUM
data_dbp$dbp <- data_dbp$VALUENUM
data_sbp$sbp <- data_sbp$VALUENUM

D=c()
D$HADM_ID <- cohort$HADM_ID
Dhr <- data_hr %>% select(HADM_ID, HR)
D1 <- left_join(as.data.frame(D), as.data.frame(Dhr), by = "HADM_ID")

Dcreatinine <- data_creatinine %>% select(HADM_ID, creatinine)
D2 <- left_join(as.data.frame(D1), as.data.frame(Dcreatinine), by = "HADM_ID")

Dhematocrit <- data_hematocrit %>% select(HADM_ID, hematocrit)
D3 <- left_join(as.data.frame(D2), as.data.frame(Dhematocrit), by = "HADM_ID")

Dgcs <- data_GCS %>% select(HADM_ID, gcs)
D4 <- left_join(as.data.frame(D3), as.data.frame(Dgcs), by = 'HADM_ID')

Dpotasium <- data_potassium %>% select(HADM_ID, potassium)
D5 <- left_join(as.data.frame(D4), as.data.frame(Dpotasium), by = 'HADM_ID')

Dsodium <- data_sodium  %>% select(HADM_ID, sodium)
D6 <- left_join(as.data.frame(D5), as.data.frame(Dsodium), by = 'HADM_ID')

Ddbp <- data_dbp  %>% select(HADM_ID, dbp)
D7 <- left_join(as.data.frame(D6), as.data.frame(Ddbp), by = 'HADM_ID')

Dsbp <- data_sbp  %>% select(HADM_ID, sbp)
D8 <- left_join(as.data.frame(D7), as.data.frame(Dsbp), by = 'HADM_ID')

#merge newly pulled data with the cohort
cohort_apache_dat <- as.data.frame(merge(cohort, D8, by='HADM_ID'))
cohort_apache_dat <- cbind(HADM_ID=c(cohort_apache$HADM_ID), SUBJECT_ID=c(cohort_apache$SUBJECT_ID), age=c(cohort_apache$AGE), hr=c(cohort_apache$HR), creatinine=c(cohort_apache$creatinine), hematocrit=c(cohort_apache$hematocrit), gcs=c(cohort_apache$gcs), potassium=c(cohort_apache$potassium), sodium=c(cohort_apache$sodium), dbp=c(cohort_apache$dbp), sbp=c(cohort_apache$sbp), APACHE_HF=c(cohort_apache$APACHE_HF))
```

##Feature Visualizations
```{r message=F, warning=F}
#Histograms of main variables
  #age
ggplot(data=cohort_apache, aes(cohort_apache$age)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #hr
ggplot(data=cohort_apache, aes(cohort_apache$hr)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #creatinine
ggplot(data=cohort_apache, aes(cohort_apache$creatinine)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #hematocrit
ggplot(data=cohort_apache, aes(cohort_apache$hematocrit)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #gcs
ggplot(data=cohort_apache, aes(cohort_apache$gcs)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #potassium
ggplot(data=cohort_apache, aes(cohort_apache$potassium)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #sodium
ggplot(data=cohort_apache, aes(cohort_apache$sodium)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #dbp
ggplot(data=cohort_apache, aes(cohort_apache$dbp)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #sbp
ggplot(data=cohort_apache, aes(cohort_apache$sbp)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
```

#Data Manipulation
##Feature Selection
Features were selected based on the variables needed to calculate the APACHE scores.  These features included age, heart rate, creatinine, hematocrit, Glasgo comma score, potassium, sodium, and mean arterial blood pressure.  Because MAP was not recorded for a large percentage of the cohort patients, we instead pulled data on diastolic blood pressure and systolic blood pressure and calculated MAP ([MDcalc.com, accessed 2019](https://www.mdcalc.com/mean-arterial-pressure-map#evidence)) from these components using the equation:
$$MAP=\frac{SBP+2DBP}{3}$$
Using this equation we can calculate  value of MAP even for patients who didn't have MAP directly recorded.

##Imputation
 
###Data Missingnes
As previousely stated, we found that certain features exhibited a high degree of missingness which necessitated using imputation to allow score calculations for all patients in the cohort.  The two features with the most missingness were SBP and DBP.  Although there was at least some missingness in every feature, it was generally not very high.
```{r message=F, warning=F}
#missingness analysis: number of missing values for each column
sum(is.na(cohort_apache[,4])) #hr
sum(is.na(cohort_apache[,5])) #creatinine
sum(is.na(cohort_apache[,6])) #hematocrit
sum(is.na(cohort_apache[,7])) #gcs
sum(is.na(cohort_apache[,8])) #potasium
sum(is.na(cohort_apache[,9])) #sodium
sum(is.na(cohort_apache[,10])) #dbp
sum(is.na(cohort_apache[,11])) #sbp
```
###Imputation Code
The code used to perform multiple imutation on the dataset was:
```{r message=F, warning=F}
#multiple imputation for missing data
cohort_apache_comp_missing <- as.data.frame(cohort_apache)
imputations <-mi(cohort_apache_comp_missing, n.iter=5, n.chains=1, verbose=TRUE)
cohort_apache_imputed <- complete(imputations, 1)

cohort_apache_imputed <-cohort_apache_imputed [,1:12]

kable(head(cohort_apache_imputed), caption="Sample of imputed data") %>%
  kable_styling(bootstrap_options='striped')
```
###Imputation Comparison: Features
To check wether imputation significantly changed the distribution of each feature we can compare the distribution before and after imputation for all features individually:
```{r message=F, warning=F}
#hr
hist(cohort_apache$hr, col=rgb(1,0,0,0.5),xlim=c(0,150), ylim=c(0,1500), main='Heart Rate Before/After Imputation', xlab='Heart Rate')
hist(cohort_apache_imputed$hr, col=rgb(0,0,1,0.5), add=T)
box()

#creatinine
hist(cohort_apache$creatinine, col=rgb(1,0,0,0.5),xlim=c(0,50), ylim=c(0,6000), main='Creatinine Before/After Imputation', xlab='Creatinine')
hist(cohort_apache_imputed$creatinine, col=rgb(0,0,1,0.5), add=T)
box()

#Hematocrit
hist(cohort_apache$hematocrit, col=rgb(1,0,0,0.5),xlim=c(0,60), ylim=c(0,2500), main='Hematocrit Before/After Imputation', xlab='Hematocrit')
hist(cohort_apache_imputed$hematocrit, col=rgb(0,0,1,0.5), add=T)
box()

#GCS
hist(cohort_apache$gcs, col=rgb(1,0,0,0.5),xlim=c(0,20), ylim=c(0,5000), main='GCS Before/After Imputation', xlab='GCS')
hist(cohort_apache_imputed$gcs, col=rgb(0,0,1,0.5), add=T)
box()

#Potassium
hist(cohort_apache$potassium, col=rgb(1,0,0,0.5),xlim=c(0,150), ylim=c(0,6000), main='Potassium Before/After Imputation', xlab='Potassium')
hist(cohort_apache_imputed$potassium, col=rgb(0,0,1,0.5), add=T)
box()

#Sodium
hist(cohort_apache$sodium, col=rgb(1,0,0,0.5),xlim=c(0,170), ylim=c(0,5000), main='Sodium Before/After Imputation', xlab='Sodium')
hist(cohort_apache_imputed$sodium, col=rgb(0,0,1,0.5), add=T)
box()

#dbp
hist(cohort_apache$dbp, col=rgb(1,0,0,0.5),xlim=c(0,150), ylim=c(0,2000), main='DBP Before/After Imputation', xlab='DBP')
hist(cohort_apache_imputed$dbp, col=rgb(0,0,1,0.5), add=T)
box()

#sbp
hist(cohort_apache$sbp, col=rgb(1,0,0,0.5),xlim=c(0,250), ylim=c(0,1000), main='SBP Before/After Imputation', xlab='SBP')
hist(cohort_apache_imputed$sbp, col=rgb(0,0,1,0.5), add=T)
box()
```

#Scoring
##Discussion of the APACHE-HF Scoring System
The APACHE-HF scoring system is used to predict adverse outcomes in patients diagnosed with acute heart failure ([Okazaki, H et all](https://www.sciencedirect.com/science/article/pii/S0914508714000951)).
Essentially, the higher the score is for a given patient the more likely they are to experience adverse outcomes.  The scores are calculated based on seven distinct variables: mean arterial pressure (in our case this is calculated from systolic and diastolic blood pressure) heart rate, sodium, potassium, creatinine, hematocrit,Glasgow coma score, and age.
Every patient starts with an initial score of zero and then points are added based on various criteria.  The rules for adding points to the scoring system are:

1. *MAP*: Add one point to the scoring system if MAP is less than or equal to 90.
2. *Heart Rate*: Add one point to the scoring system if heart rate is less than or equal to 110.
3. *Sodium*: Add one point to the scoring system if sodium is less than or equal to 137.
4. *Potassium*: Add one point to the scoring system if potassium is greater than or equal to 4.9.
5. *Creatinine*: Add one point to the scoring system if creatinine is greater than or equal to 1.48.
6. *Hematocrit*: Add one point to the scoring system if hematocrit is less than or equal to 36.9.
7. *Glasgow Coma Score*: Add one point to the scoring system if GCS is less than or equal to 13.
8. *Age*: Add one point to the scoring system if age is greater than or equal to 72.

This process will assign each patient a clinically meaningfull score between 0 and 8 to asses  their risk for adverse events based on their heart failure diagnosis and available lab data.
##Calculation Code
Calculations of scores were made according to the published APACHE score documentations, which listed out rules for determinimg a patient's score based on the various features.
```{r message=F, warning=F}
#calculate the APACHE-HF score for complete cases (no missingness):
cohort_apache_comp = as.data.frame(cohort_apache)
cohort_apache_comp$APACHE_HF <- rep(x=0, times = nrow(cohort_apache_comp))
cohort_apache_comp$APACHE_HF <- ifelse((cohort_apache_comp$sbp+2*cohort_apache_comp$dbp)/3<=90, yes = 1, no = cohort_apache_comp$APACHE_HF)
cohort_apache_comp$APACHE_HF <- ifelse(cohort_apache_comp$hr<=110, yes = cohort_apache_comp$APACHE_HF+1, no = cohort_apache_comp$APACHE_HF)
cohort_apache_comp$APACHE_HF <- ifelse(cohort_apache_comp$sodium<=137, yes = cohort_apache_comp$APACHE_HF+1, no = cohort_apache_comp$APACHE_HF)
cohort_apache_comp$APACHE_HF <- ifelse(cohort_apache_comp$potassium>=4.9, yes = cohort_apache_comp$APACHE_HF+1, no = cohort_apache_comp$APACHE_HF)
cohort_apache_comp$APACHE_HF <- ifelse(cohort_apache_comp$creatinine>=1.48, yes = cohort_apache_comp$APACHE_HF+1, no = cohort_apache_comp$APACHE_HF)
cohort_apache_comp$APACHE_HF <- ifelse(cohort_apache_comp$hematocrit<=36.9, yes = cohort_apache_comp$APACHE_HF+1, no = cohort_apache_comp$APACHE_HF)
cohort_apache_comp$APACHE_HF <- ifelse(cohort_apache_comp$gcs<=13, yes = cohort_apache_comp$APACHE_HF+1, no = cohort_apache_comp$APACHE_HF)
cohort_apache_comp$APACHE_HF <- ifelse(cohort_apache_comp$age>=72, yes = cohort_apache_comp$APACHE_HF+1, no = cohort_apache_comp$APACHE_HF)

APACHE_HF = cohort_apache_comp$APACHE_HF

#calculate the APACHE-HF score after imputation (all cases, missing data imputed): 
cohort_apache_imputed$APACHE_HF <- rep(x=0, times = nrow(cohort_apache_imputed))
cohort_apache_imputed$APACHE_HF <- ifelse((cohort_apache_imputed$sbp+2*cohort_apache_imputed$dbp)/3<=90, yes = 1, no = cohort_apache_imputed$APACHE_HF)
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$hr<=110, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$sodium<=137, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$potassium>=4.9, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$creatinine>=1.48, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$hematocrit<=36.9, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$gcs<=13, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$age>=72, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)


APACHE_HF_imputed = cohort_apache_imputed$APACHE_HF
```
##Imputation Comparison: Scores
Next we need to determine if the imputed data points altered the overall distribution of the APACHE scores calculated to cohort patients.  To do so, we compare the histogram of the scores based only on patients with complete data with the histogram of scores from all patients after imputation is used to infer values for missing data points:
```{r message=F, warning=F}
h1 = hist(APACHE_HF, col=rgb(1,0,0,0.5),xlim=c(0,8), ylim=c(0,2500), main='Scores Before Imputation', xlab='APACHE Score')
print(h1)
```
```{r message=F, warning=F}
h2 = hist(APACHE_HF_imputed, col=rgb(0,0,1,0.5),xlim=c(0,8), ylim=c(0,2500), main='Scores After Imputation', xlab='APACHE Score')
print(h2)
```
```{r message=F, warning=F}
hist(APACHE_HF, col=rgb(1,0,0,0.5),xlim=c(0,8), ylim=c(0,2500), main='Scores Before/After Imputation', xlab='APACHE Score')
hist(APACHE_HF_imputed, col=rgb(0,0,1,0.5), add=T)
box()
```

#Risk Analysis
##Mortality Calculation
Based on the information available for each patient we can calculate the time until mortality by subtracting time of discharge from hospital from the patient's time of mortaliaty.  

For patient i:
$$T_{Mortality,i}=T_{Death,i}-T_{Discharge,i}$$
To calculate time to mortality for each patient in the cohort:

```{r message=F, warning=F}
cohort_for_mortality <- cohort
cohort_for_mortality$Score <- APACHE_HF_imputed
cohort_for_mortality <- cohort_for_mortality[!is.na(cohort_for_mortality$DOD),]

time_to_mortality = as.numeric(difftime(as.Date(cohort_for_mortality$DOD, format = "%m/%d/%Y"), as.Date(cohort_for_mortality$DISCHTIME.x, format = "%m/%d/%Y"), units = c('days')))
hist(time_to_mortality)
```

##Score vs Mortlity (Binary)
Mortality can be treated as a binary variablt, determining patients who survive for a certin amount of time after discharge, to model the relationship between APACHE score and time of mortality.

We considered seven different time to mortality ranges, separating patients into those who survived for 0 days after discharge, 2 days after discharge, 2 weeks after discharge, 6 months after discharge, 1 year after discharge, 2 years after discharge, and over 2 years after discharge.

For each category we can then determine the mean APACHE score for patients in each of the catogories.
```{r message=F, warning=F}
HADM_ID = cohort_for_mortality$HADM_ID
Score = cohort_for_mortality$Score
Mortality = time_to_mortality
cohort_apache_mortality = cbind(HADM_ID, Score, Mortality)
cohort_apache_mortality_0days = subset(cohort_apache_mortality, Mortality == 0)
cohort_apache_mortality_2days = subset(cohort_apache_mortality, Mortality<=2 & Mortality>0)
cohort_apache_mortality_2weeks = subset(cohort_apache_mortality, Mortality<=14 & Mortality>2)
cohort_apache_mortality_6months = subset(cohort_apache_mortality, Mortality<=182.5 & Mortality>14)
cohort_apache_mortality_1year = subset(cohort_apache_mortality, Mortality<=365 & Mortality>182.5)
cohort_apache_mortality_2years = subset(cohort_apache_mortality, Mortality<=730 & Mortality>365)
cohort_apache_mortality_over2years = subset(cohort_apache_mortality, Mortality>730)

mean_score = c(mean(cohort_apache_mortality_0days[,2]), mean(cohort_apache_mortality_2days[,2]),
               mean(cohort_apache_mortality_2weeks[,2]), mean(cohort_apache_mortality_6months[,2]), 
               mean(cohort_apache_mortality_1year[,2]), mean(cohort_apache_mortality_2years[,2]), 
               mean(cohort_apache_mortality_over2years[,2]))
lower_score = c(t.test(cohort_apache_mortality_0days[,2])$"conf.int"[1],
                t.test(cohort_apache_mortality_2days[,2])$"conf.int"[1],
                t.test(cohort_apache_mortality_2weeks[,2])$"conf.int"[1],
                t.test(cohort_apache_mortality_6months[,2])$"conf.int"[1],
                t.test(cohort_apache_mortality_1year[,2])$"conf.int"[1],
                t.test(cohort_apache_mortality_2years[,2])$"conf.int"[1],
                t.test(cohort_apache_mortality_over2years[,2])$"conf.int"[1])
upper_score = c(t.test(cohort_apache_mortality_0days[,2])$"conf.int"[2],
                t.test(cohort_apache_mortality_2days[,2])$"conf.int"[2],
                t.test(cohort_apache_mortality_2weeks[,2])$"conf.int"[2],
                t.test(cohort_apache_mortality_6months[,2])$"conf.int"[2],
                t.test(cohort_apache_mortality_1year[,2])$"conf.int"[2],
                t.test(cohort_apache_mortality_2years[,2])$"conf.int"[2],
                t.test(cohort_apache_mortality_over2years[,2])$"conf.int"[2])
bin_size = c('0 days', '2 days', '2 weeks', '6 months', '1 year', '2 years', 'over 2 years')
bin_size2=factor(bin_size, levels = bin_size)
mean_score = round(mean_score, 3)
lower_score = round(lower_score,3)
upper_score = round(upper_score,3)
mortality_CI = as.data.frame(cbind(bin_size2, mean_score, lower_score, upper_score))

#visualize for different bin sizes
ggplot() + 
  geom_errorbar(data=mortality_CI, mapping=aes(x=bin_size, ymin=upper_score, ymax=lower_score), width=0.2, size=1, color="blue") + 
  geom_point(data=mortality_CI, mapping=aes(x=bin_size, y=mean_score), size=4, shape=21, fill="white") +
  geom_text(aes(label = round(mean_score, digits = 2)), x = c(1,2,3,4,5,6,7)+0.4, y = mean_score-0.03, vjust = -0.25) +
  ggtitle("Time to Mortality as a Discrete Variable") + xlab("Bin Size") + ylab("Mean APACHE Score")
 
```

##Score vs Mortality (Continuous)
If we treat time to mortality instead as a continuous variable, it is possible to visualize the connection between a patient's APACHE score calculated at discharge and their expected time to mortality.

For each of the possible APACHE scores (1 to 8) we can visualize the average time to mortality for patients who receive that score upon discharge:
```{r message=F, warning=F}
# calculate mean time to mortality for each score
dat=cbind(Score, time_to_mortality)
dat=as.data.frame(dat)
plotdata <- dat %>%
  group_by(Score) %>%
  summarize(mean_ttm = mean(time_to_mortality))

# plot mean ttm
ggplot(plotdata, 
       aes(x = factor(Score,
          labels = c('0','1','2','3','4','5','6','7')), 
           y = mean_ttm)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = round(mean_ttm, digits = 2)), 
            vjust = -0.25) +
  scale_y_continuous(breaks = seq(0,2000, 100)) +
  labs(title = "Mean Time to Mortality by APACHE Score", 
       subtitle = "measured in days",
       x = "",
       y = "")
```

##Survival Curve
Using time to mortality as a continuous variable, we can estimate a survival curve based on the data:
```{r message=F, warning=F}
cohort_for_survival <- merge(cohort, cohort_apache)
cohort_for_survival$Score <- APACHE_HF_imputed
cohort_for_survival <- cohort_for_survival[!is.na(cohort_for_survival$DOD),]
cohort_for_survival$mortality = time_to_mortality

# status 0 = alive at 6 months, 1 = died between discharge and 6 months
cohort_for_survival$status <- rep(x=0, times = nrow(cohort_for_survival))
cohort_for_survival$status <- ifelse(cohort_for_survival$mortality>182.5,yes=cohort_for_survival$status, no=cohort_for_survival$status+1)
pd_surv_fit <- survfit(Surv(Score,status)~1, data=cohort_for_survival)
plot(pd_surv_fit, xlab="APACHE score", ylab="Survival Probability")
summary(pd_surv_fit)
```

```{r message=F, warning=F}
pd_surv_fit2 <- survfit(Surv(hr,status)~1, data=cohort_for_survival)
ggsurvplot(pd_surv_fit2, xlab="Time in hospital")

# survival polt for score 0-2
cohort_for_survival_interval02 <- cohort_for_survival[cohort_for_survival$Score<=2,]
pd_surv_fit02 <- survfit(Surv(hr,status)~1, data=cohort_for_survival_interval02)
ggsurvplot(pd_surv_fit02,xlab="Time in hospital")
#survival plot for score 3-5
cohort_for_survival_interval35 <- cohort_for_survival[which(cohort_for_survival$Score>=3 & cohort_for_survival$Score<=5),]
pd_surv_fit35 <- survfit(Surv(hr,status)~1, data=cohort_for_survival_interval35)
ggsurvplot(pd_surv_fit35,xlab="Time in hospital")
# survival plot for score greater or equal to 6
cohort_for_survival_interval6 <- cohort_for_survival[which(cohort_for_survival$Score>=6),]
pd_surv_fit6 <- survfit(Surv(hr,status)~1, data=cohort_for_survival_interval6)
ggsurvplot(pd_surv_fit6,xlab="Time in hospital")
```

# References

Okazaki H, Shirakabe A, Hata N, Yamamoto M, Kobayashi N,
Shinada T, Tomita K, Tsurumi M, Matsushita M, Yamamoto Y,
Yokoyama S, Asai K, Shimizu W. 
New scoring system (APACHEHF) for predicting adverse outcomes in patients with 
acute heart failure: evaluation of the APACHE II and Modified APACHE II scoring systems. 
J Cardiol 2014; 64: 441-449 [PMID: 24794758 DOI: 10.1016/j.jjcc.2014.03.002]

Passantino, Andrea & Monitillo, Francesco & Iacoviello, Massimo & Scrutinio, Domenico. (2015). 
Predicting mortality in patients with acute heart failure: Role of risk scores. 
World Journal of Cardiology. 7. 902. 10.4330/wjc.v7.i12.902. 

MDCalc.com, Mean Arterial Pressure MAP, accessed 2019, https://www.mdcalc.com/mean-arterial-pressure-map#evidence

