---
title: "MIMIC III: APACHE-HF Score Calculation"
author: 
- "<h3><p>Jessica Wild & Ziyuan Sun</p><p>jesswild@umich.edu ziyuans@umich.edu</p></h3>"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: default
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
subtitle: <h2><u>Winter 2019, SOCR-MDP</u></h2>
---

**Libraries Used:**
```{r message=F, warning=F}
# Data I/O
require('RPostgreSQL')    # access MIMIC-III
require('reshape2')       # melting dataframes
require('dplyr')          # misc. organization
# Data imputation
require('mi')
# Plots and tables
require('knitr')          # knitting Rmd to HTML; kable() function
require('kableExtra')     # extra formating options for knitr tables
require('ggplot2')        # 2d plotting
```

**Random Seed Set**
```{r message=F, warning=F}
set.seed(11042019)
```

# Load Raw Data

Data needed to calculate APACHE scores was pulled directly from MIMIC and stores as a csv file. 

```{r eval=T, message=F, warning=F}
#age, hr, creatinine, hematocrit, gcs, potassium, sodium, dbp, sbp
cohort_apache <- read.csv('Data_Apache.csv')
kable(head(cohort_apache), caption="Sample of raw data") %>%
  kable_styling(bootstrap_options='striped')
```

In total we have `r nrow(cohort_apache)` admission ID's within the cohort.

# Impute Missing Values

There is quite a bit of missingness present in the data set, more in some variables than others.  To allow us to calculate the  APACHE scores for all admission ID's we need to impute plausible estimates for these missing values

```{r message=F, warning=F}
cohort_apache_comp_missing <- as.data.frame(cohort_apache)
imputations <-mi(cohort_apache_comp_missing, n.iter=5, n.chains=1, verbose=TRUE)
cohort_apache_imputed <- complete(imputations, 1)

cohort_apache_imputed <-cohort_apache_imputed [,1:11]

kable(head(cohort_apache_imputed), caption="Sample of imputed data") %>%
  kable_styling(bootstrap_options='striped')
```

# Calculate MAP from SBP & DBP

Mean arterial pressure was not recorded often enough in MIMIC for us to directly pull enough data to ensure accurate imputation.  Instead, we pulled data on systolic and diasystolic blood pressure and calculated MAP with a function.

```{r message=F, warning=F}
# function to calculate map
mean_arterial_pressure <- function(sbp,dbp){
  return((sbp+2*dbp)/3)
}

cohort_apache_imputed = as.data.frame(cohort_apache_imputed)
cohort_apache_imputed$map = mean_arterial_pressure(cohort_apache_imputed$sbp, cohort_apache_imputed$dbp)

kable(head(cohort_apache_imputed$map), caption="Sample of calculated MAP") %>%
  kable_styling(bootstrap_options='striped')
```

# Calculate APACHE-HF Scores 

```{r message=F, warning=F}
#calculate the APACHE-HF score after imputation:
cohort_apache_imputed$APACHE_HF <- rep(x=0, times = nrow(cohort_apache_imputed))
  #set up an empty vector to store each admission ID's score
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$map<=200, yes = 1, no = cohort_apache_imputed$APACHE_HF)
  #map
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$hr<=110, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
  #hr
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$sodium<=137, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
  #sodium
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$potassium>=4.9, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
  #potassium
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$creatinine>=1.48, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
  #creatinine
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$hematocrit<=36.9, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
  #hematocrit
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$gcs<=13, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
  #gcs
cohort_apache_imputed$APACHE_HF <- ifelse(cohort_apache_imputed$age>=72, yes = cohort_apache_imputed$APACHE_HF+1, no = cohort_apache_imputed$APACHE_HF)
  #age
APACHE_HF_imputed = cohort_apache_imputed$APACHE_HF

kable(head(APACHE_HF_imputed), caption="Sample of calculated APACHE-HF Scores") %>%
  kable_styling(bootstrap_options='striped')
```

#Data Visualization
```{r message=F, warning=F}
#Histograms of main variables
  #age
ggplot(data=cohort_apache_imputed, aes(cohort_apache_imputed$age)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #hr
ggplot(data=cohort_apache_imputed, aes(cohort_apache_imputed$hr)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #creatinine
ggplot(data=cohort_apache_imputed, aes(cohort_apache_imputed$creatinine)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #hematocrit
ggplot(data=cohort_apache_imputed, aes(cohort_apache_imputed$hematocrit)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #gcs
ggplot(data=cohort_apache_imputed, aes(cohort_apache_imputed$gcs)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #potassium
ggplot(data=cohort_apache_imputed, aes(cohort_apache_imputed$potassium)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #sodium
ggplot(data=cohort_apache_imputed, aes(cohort_apache_imputed$sodium)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
  #map
ggplot(data=cohort_apache_imputed, aes(cohort_apache_imputed$map)) + geom_histogram(breaks=seq(0, 200, by =2), col="red", aes(fill=..count..))
#Pie Chart of APACHE Scores
class = c('1','2','3','4','5','6','7','8')
prop = c(sum(APACHE_HF_imputed==1),sum(APACHE_HF_imputed==2),sum(APACHE_HF_imputed==3),sum(APACHE_HF_imputed==4),sum(APACHE_HF_imputed==5),sum(APACHE_HF_imputed==6),sum(APACHE_HF_imputed==7),sum(APACHE_HF_imputed==8))
prop = 100*prop/sum(prop)
data = as.data.frame(cbind(prop,class))
ggplot(, aes(x = "", y = prop, fill = class)) +
  geom_bar(width = 1, stat = "identity", color = "white") +
  coord_polar("y", start = 0)+
  theme_void()
```
