 ---
title: "MIMIC-III OPTIME-CHF Score Calculation Demo"
author: 
- "<h3><p>Audrey Zhang</p><p>audzhang@umich.edu</p></h3>"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: default
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
subtitle: <h2><u>Winter 2019, SOCR-MDP</u></h2>
---

**Libraries Used:**
```{r message=F, warning=F}
# Data I/O
library('RPostgreSQL')    # access MIMIC-III
library('reshape2')       # melting dataframes
library('dplyr')          # misc. organization

# Data preparation
library('psych')          # descriptive stats

# Plots and tables
library('knitr')          # knitting Rmd to HTML; kable() function
library('kableExtra')     # extra formating options for knitr tables
library('ggplot2')        # 2d plotting
library('ggpubr')         # extra formatting options for ggplot
library('gtools')

```

**Random Seed Set**
```{r message=F, warning=F}
set.seed(123456)
```

# Introduction

First, load in the CSV containing our cohort 

```{r eval=T, message=F, warning=F}
#age, cbv disease, dementia, COPD
cohort_data <- read.csv('new_cohort.csv')

kable(head(cohort_data), caption="Sample of cohort data") %>%
  kable_styling(bootstrap_options='striped')
```

###BUN + creatinine + hemoglobin + resp
```{r eval=T, message=F, warning=F}
ccbhr <- read.csv('cirr_crea_bun_hemo_resp.csv')

# Add to cohort_data object
cohort_data <- merge(cohort_data, ccbhr[,c('HADM_ID', 'CREATININE', 'HEMOGLOBIN', 'BUN', 'RESPIRATORY_RATE')], by=c('HADM_ID'), all=T)

colnames(cohort_data) <- c(colnames(cohort_data)[1:(length(cohort_data)-1)], 'RESP_RATE')

# cohort_data = cohort_data[!duplicated(cohort_data$subject_id),]

```

###cirr
```{r eval=T, message=F, warning=F}
cirr <- read.csv('cirr.csv')

# Add to cohort_data object
cohort_data <- merge(cohort_data, cirr[,c('HADM_ID', 'cirrhosis')], by=c('HADM_ID'), all=T)

colnames(cohort_data) <- c(colnames(cohort_data)[1:(length(cohort_data)-1)], 'cirrhosis')
```


###sbp + sodium
```{r eval=T, message=F, warning=F}
ss <- read.csv('sodium_sbp.csv')

# Add to cohort_data object
cohort_data <- merge(cohort_data, ss[,c('HADM_ID', 'SBP')], by=c('HADM_ID'), all=T)

colnames(cohort_data) <- c(colnames(cohort_data)[1:(length(cohort_data)-1)], 'SBP')

```


###depression
```{r eval=T, message=F, warning=F}
liver <- read.csv('liver_subject_icds.csv')
reac <- read.csv('reative_airway_disease_subject_ids.csv')
depre <- read.csv('depression.csv')

# Add to cohort_data object
cohort_data <- merge(cohort_data, liver[,c('SUBJECT_ID', 'liver_disease')], by=c('SUBJECT_ID'), all=T)
colnames(cohort_data) <- c(colnames(cohort_data)[1:(length(cohort_data)-1)], 'LIVER')

cohort_data <- merge(cohort_data, reac[,c('SUBJECT_ID', 'airway_disease')], by=c('SUBJECT_ID'), all=T)
colnames(cohort_data) <- c(colnames(cohort_data)[1:(length(cohort_data)-1)], 'REAC_AIR')

cohort_data <- merge(cohort_data, depre[,c('HADM_ID', 'depression')], by=c('HADM_ID'), all=T)
colnames(cohort_data) <- c(colnames(cohort_data)[1:(length(cohort_data)-1)], 'DEPRESSION')


```


Lastly, we'll write the output to CSV for part two

```{r eval=T, message=F, warning=F}
#take out duplicate hadmids
cohort_data <- cohort_data[!duplicated(cohort_data[,c('HADM_ID')]),]
write.csv(cohort_data, file='effect_predictions_final.csv', row.names=F)
```

## Remove missing variables

We'll also choose to remove rows that contain missing variables - this makes visualization and exploratory data analysis easier, while unfortunately reducing our overall sample size.

```{r message=F, warning=F}
cohort_data_complete = cohort_data[complete.cases(cohort_data),]
```

After removing rows with missing values, we're left with `r nrow(cohort_data)` subjects. This is still plenty for our analysis.

## Calculate EFFECT Scores 

```{r message=F, warning=F}
#age, cbv disease, dementia, COPD, BUN + cirrhosis + hemoglobin + resp, SBP, SODIUM

for (i in 1:nrow(cohort_data_complete)){
  score <- 0
  #age
  score <- score + cohort_data_complete$AGE[i]
  if( cohort_data_complete$CV_Disease[i] == 1){
    score <- score + 10
  }
  if( cohort_data_complete$Dementia[i] == 1){
    score <- score + 10
  }
  if (cohort_data_complete$COPD[i] == 1){
    score <- score + 20
  }
  # do nothing for hemoglobin i guess
  if(cohort_data_complete$BUN[i] <= 60){
    score <- score + cohort_data_complete$BUN[i]
  }
  if(cohort_data_complete$RESP_RATE[i] >= 20 && cohort_data_complete$RESP_RATE[i] <= 45){
    score <- score + cohort_data_complete$RESP_RATE[i]
  }
  if( cohort_data_complete$cirrhosis[i] == 1){
    score <- score + 25
  }
  # sbp
  if(cohort_data_complete$SBP[i] >= 180){
    score <- score - 60
  } else if(cohort_data_complete$SBP[i] >= 160){
    score <- score - 55
  } else if(cohort_data_complete$SBP[i] >= 140){
    score <- score - 50
  } else if(cohort_data_complete$SBP[i] >= 120){
    score <- score- 45
  } else if(cohort_data_complete$SBP[i] >= 100){
    score <- score - 40
  } else if(cohort_data_complete$SBP[i] >= 90){
    score <- score - 35
  } else if(cohort_data_complete$SBP[i] < 90){
    score <- score - 30
  }
  if(cohort_data_complete$SODIUM[i] < 136){
   score <- score + 10
  }
  cohort_data_complete$EFFECT[i] <- score
}
```


## Calculate OPTIMIZE Scores 
```{r message=F, warning=F}
#age, cbv disease, dementia, COPD, BUN + cirrhosis + hemoglobin + resp, SBP, SODIUM
cohort_data_complete <- read.csv('DATA.csv')

for (i in 1:nrow(cohort_data_complete)){
  score <- 0
  #age
  if( cohort_data_complete$AGE[i] < 30){
    score <- score + 0
  } else if( cohort_data_complete$AGE[i] < 40){
    score <- score + 2
  }  else if( cohort_data_complete$AGE[i] < 50){
    score <- score + 5
  }  else if( cohort_data_complete$AGE[i] < 60){
    score <- score + 7
  }  else if( cohort_data_complete$AGE[i] < 70){
    score <- score + 10
  }  else if( cohort_data_complete$AGE[i] < 80){
    score <- score + 12
  }  else if( cohort_data_complete$AGE[i] < 90){
    score <- score + 15 
  } else if( cohort_data_complete$AGE[i] >= 90){
    score <- score + 17 
  }  
  # weight, kg
  if( cohort_data_complete$WEIGHT[i] < 70){
    score <- score + 9
  } else if( cohort_data_complete$WEIGHT[i] < 90){
    score <- score + 7
  } else if( cohort_data_complete$WEIGHT[i] < 110){
    score <- score + 5
  } else if( cohort_data_complete$WEIGHT[i] < 130){
    score <- score + 3
  } else if( cohort_data_complete$WEIGHT[i] >= 130){
    score <- score + 2
  }
  # sbp
  if(cohort_data_complete$SBP[i] < 90){
    score <- score + 24
  } else if(cohort_data_complete$SBP[i] < 110){
    score <- score + 20
  } else if(cohort_data_complete$SBP[i] < 130){
    score <- score + 17
  } else if(cohort_data_complete$SBP[i] < 150){
    score <- score + 13
  } else if(cohort_data_complete$SBP[i] < 170){
    score <- score + 11
  } else if(cohort_data_complete$SBP[i] < 190){
    score <- score + 9
  } else if(cohort_data_complete$SBP[i] < 210){
    score <- score +8
  } else if(cohort_data_complete$SBP[i] < 230){
    score <- score +6
  } else if(cohort_data_complete$SBP[i] < 250){
    score <- score +4
  } else if(cohort_data_complete$SBP[i] < 270){
    score <- score +2
  } else if(cohort_data_complete$SBP[i] >= 270){
    score <- score + 0
  } 
  
  #sodium
  if(cohort_data_complete$SODIUM[i] < 112.5){
   score <- score + 12
  } else if(cohort_data_complete$SODIUM[i] < 117.5){
   score <- score + 10
  } else if(cohort_data_complete$SODIUM[i] < 122.5){
   score <- score + 8
  } else if(cohort_data_complete$SODIUM[i] < 127.5){
   score <- score + 6
  } else if(cohort_data_complete$SODIUM[i] < 132.5){
   score <- score + 4
  } else if(cohort_data_complete$SODIUM[i] < 137.5){
   score <- score + 2
  } else if(cohort_data_complete$SODIUM[i] >= 137.5 ){
   score <- score + 0
  }
  #creatinine 
  if (cohort_data_complete$CREATININE[i] < 0.5){
    score <- score + 0
  } else if (cohort_data_complete$CREATININE[i] < 1.5){
    score <- score + 5
  } else if (cohort_data_complete$CREATININE[i] < 2.5){
    score <- score + 9
  } else if (cohort_data_complete$CREATININE[i] < 3.5){
    score <- score + 14
  } else if (cohort_data_complete$CREATININE[i] >= 3.5){
    score <- score + 19
  }
  
  # history of liver disease 
  if(cohort_data_complete$LIVER[i] == 1){
    score <- score + 8
  }
  
  #history of reactive airway disease 
  if( cohort_data_complete$REAC_AIR[i] == 1){
    score <- score + 4
  }
  #history of depression
  if( cohort_data_complete$DEPRESSION[i] == 1){
    score <- score + 4
  }
  
  cohort_data_complete$OPTIMIZE[i] <- score
}
```


```{r message=F, warning=F}
# Avg 30 day mortaility per quantiles

# make quantiles
cohort_data_complete$quantiles <- quantcut(cohort_data_complete$EFFECT, q = seq(0, 1, by=.2), na.rm=T)
# get everyone who died within 30 days
cohort_30_days <- cohort_data_complete[cohort_data_complete$DOD_30days==1,]

# get average mortality per quantile
quantiles_30_day <- aggregate.data.frame(cohort_30_days$DISCHtoDOD, by=list(quantiles=cohort_30_days$quantiles), FUN=mean)

cohort_data_complete <- cohort_data_complete[cohort_data_complete$DISCHtoDOD > 0, ]

# graph it
fig <- ggplot(data=quantiles_30_day, aes(x=quantiles, y=x, group=1)) +
  geom_line() + geom_point() +
  labs(y = "Average number of days lived past discharge", x = "Risk Score Quantiles") +
  ggtitle("Average number of days past discharge per quantile")
fig
```

##IMPUTATION
```{r message=F, warning=F}
library(Amelia)
sim_data.df <- read.csv(file='DATA_MASTER.csv', header=TRUE)
dim(sim_data.df)
amelia.out <- amelia(sim_data.df, m = 5)
amelia.out
amelia.imputed.5 <- amelia.out$imputations[[5]]


imputed_scores <- read.csv(file='DATA_IMPUTE_EFFECT_OPTIME_OPTIMIZE.csv', header=TRUE)
normal_scores <- read.csv(file='DATA_effect_optime_optimize.csv', header=TRUE)

```

